### 2025-03-18 chapter/paper.tex
add skeleton for latex file for chapter

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}

% need to have the pythonhighlight.sty in directory
\usepackage{pythonhighlight}

% local definitions
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}

\begin{document}

\title{Demographic inference using the SFS with Moments and Demes}
\author[1]{Nick Collier}
%\author[2]{Simon Gravel(?)}
\author[1,*]{Aaron P. Ragsdale}
\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison}
%\affil[2]{McGill University}
\affil[*]{apragsdale@wisc.edu}
\maketitle

\abstract{
Placeholder
}

\section{Introduction}

What moments is, what type of data it uses, general framework for inference
\begin{enumerate}
    \item The SFS -- how its shape is sensitive to population,
        selection and genome biological parameters
    \item Demographic inference from the SFS
    \item Extensions: inference of selection, mutation rates, etc
    \item Extensions: Multi-allele and multi-locus models
\end{enumerate}

\section{Interface with Demes}

\begin{enumerate}
    \item Manipulating and plotting Demes-specified demographic models
    \item Computing statistics given a Demes model
    \item Inferring history on a demes model, including specifying parameters to be fit
\end{enumerate}

\section{Examples}

Two examples, one using simulated data with known parameters, which we try to
reinfer. And another with actual data (humans + neanderthal). We briefly
describe these, highlighting the high-level components of the fits. For
detailed information for each, including managing data, specifying models,
performing inference, and visualizing results, we refer readers to the GitHub
(\url{}), which is maintained with up-to-date versioning.

\subsection{Inferring parameters in a simulated isolation-with-migration model}

A python snippet:
\begin{python}
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{python}
or
\begin{python}
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{python}

This was in discussion of \citet{kelleher2016efficient}.

\subsection{Inferring human-Neanderthal demographic parameters}

\begin{python}
def myFunc():
    pass
\end{python}

\section{Considerations and caveats}

Incluce here general ideas about the strengths and weaknesses of various approaches
in population genetic inference, including when using the SFS.

Challenges in finding local/global optima. Challenges in exploring parameter space.

\bibliographystyle{plainnat}
\bibliography{paper}

\clearpage

\appendix
\renewcommand{\thesection}{A\arabic{section}}
\renewcommand{\theequation}{A\arabic{equation}}
\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\setcounter{equation}{0}
\setcounter{table}{0}

%\section{Appendix} \label{appendix}
\end{document}

### 2025-03-18 chapter/paper.tex
get at least a bit of writing going

@@
 
+\newcommand{\moments}{\texttt{moments}\xspace}
+
 \begin{document}
@@
 
-\abstract{
+\begin{abstract}
     Placeholder
-}
+\end{abstract}
 
 \section{Introduction}
+
+The genetic composition within a sample of individuals is shaped by their
+genome biology and evolutionary history. Variation resulting from this history
+is fully described by the ancestral relationships among samples at each locus
+in the genome and how they change along a chromosome due to recombination (that
+is, information stored in the Ancestral Recombination Graph [cites]). However,
+the ARG can be large and unwieldy, and methods for reconstructing history
+directly from the ARG, while showing promise [cite], are in their infancy and
+so far limited in application and scalability [cite]. Instead, evolutionary
+inference using informative summaries provides a tractable and powerful
+alternative for learning parameters of population history, natural selection and
+genome biology.
+
+One such summary that has seen wide use is the site frequency spectrum (SFS),
+which stores the counts (observed or expected) of alleles carried by a given
+number of genomes in a set of samples. Like any summary of the data, the SFS
+discards information relative to the ARG -- in this case, loci are treated
+independently so that haplotypic information is lost. Nonetheless, a great deal
+can be learned from the
 
@@
 
-Incluce here general ideas about the strengths and weaknesses of various approaches
-in population genetic inference, including when using the SFS.
-
-Challenges in finding local/global optima. Challenges in exploring parameter space.
+\begin{enumerate}
+    \item Incluce here general ideas about the strengths and weaknesses of
+        various approaches in population genetic inference, including when
+        using the SFS.
+    \item Challenges in finding local/global optima.
+    \item Challenges in exploring parameter space.
+    \item Gene dense vs gene sparse genomic architectures among species.
+\end{enumerate}
 

### 2025-03-18 example1/fit_misspec_model.py
changes to example 1

@@ -1,0 +1,46 @@
+import moments
+import gzip
+import pickle
+
+# Load the data
+data = moments.Spectrum.from_file("data/data.fs")
+
+# Specify model paths
+g_in = "model.misspec.yaml"
+options = "model.misspec.options.yaml"
+g_out = "model.misspec.out.yaml"
+
+# Parameters from simulated data
+u = 1e-8
+num_reps = 500
+L = 1e6
+U = u * L * num_reps
+
+
+# Run fit
+moments.Demes.Inference.optimize(
+    g_in,
+    options,
+    data,
+    verbose=1,
+    method="fmin",
+    perturb=1,
+    uL=U,
+    output=g_out,
+    overwrite=True
+)
+
+# Compute CIs
+with gzip.open("data/bootstrapped_data.pkl.gz", "rb") as fin:
+    bs_data = pickle.load(fin)
+
+moments.Demes.Inference.uncerts(
+    g_out,
+    options,
+    fs,
+    bootstraps=bs_data,
+    uL=U,
+    bootstraps_uL=[U] * len(bs_data),
+    verbose=1,
+    method="GIM",
+)

### 2025-03-18 example1/plot_misspec_model.py

@@ -1,0 +1,25 @@
+import demes, demesdraw
+import matplotlib.pylab as plt
+import numpy as np
+
+g = demes.load("model.yaml")
+g2 = demes.load("model.misspec.out.yaml")
+
+fig = plt.figure(figsize=(10, 4))
+ax1 = plt.subplot(1, 2, 1)
+ax2 = plt.subplot(1, 2, 2)
+
+demesdraw.tubes(g, ax=ax1)
+demesdraw.tubes(g2, ax=ax2, inf_ratio=0.55)
+
+w = ax1.get_xlim()[1] - ax1.get_xlim()[0]
+m2 = np.mean(ax2.get_xlim())
+ax2.set_xlim((m2 - w/2, m2 + w/2))
+ax2.set_ylim(top=ax1.get_ylim()[1])
+
+ax1.set_title("Simulated model")
+ax2.set_title("Inferred model")
+
+fig.tight_layout()
+fig.savefig("model.misspec.fit.png")
+plt.show()

### 2025-03-18 example1/simulate_data.py

@@ -1,0 +1,73 @@
+import demes
+import msprime
+import numpy as np
+import moments
+import pickle
+import gzip
+
+_pop_ids = ["popA", "popB"]
+
+def run_sims():
+    """
+    Returns a list of frequency spectra, over some number of replicates.
+    """
+    g = demes.load("model.yaml")
+
+    # sample sizes of 30 diploids from each population
+    n = 30
+    sample_sets = [msprime.SampleSet(n, _pop_ids[0]), msprime.SampleSet(n, _pop_ids[1])]
+    demog = msprime.Demography.from_demes(g)
+
+    # simulate 500 1Mb regions
+    L = 1e6
+    u = 1e-8
+    r = 1e-8
+    num_reps = 500
+
+    tss = msprime.sim_ancestry(
+        samples=sample_sets,
+        demography=demog,
+        sequence_length=L,
+        recombination_rate=r,
+        num_replicates=num_reps,
+        random_seed=42,
+    )
+
+    # get list of replicate spectra
+    spectra = np.zeros((num_reps, 2*n+1, 2*n+1))
+    for i, ts in enumerate(tss):
+        mts = msprime.sim_mutations(ts, rate=u, random_seed=i+13)
+        fs_rep = mts.allele_frequency_spectrum(
+            sample_sets=[range(2*n), range(2*n, 4*n)],
+            polarised=True,
+            span_normalise=False
+        )
+        spectra[i] = fs_rep
+    return spectra
+
+
+def bootstrap_spectra(spectra):
+    """
+    TODO: complete this docstring
+    """
+    num_reps = len(spectra)
+    bs_spectra = []
+    for _ in range(num_reps):
+        choices = np.random.choice(num_reps, num_reps, replace=True)
+        fs_rep = np.sum(spectra[choices], axis=0)
+        bs_spectra.append(moments.Spectrum(fs_rep, pop_ids=_pop_ids))
+    return bs_spectra
+
+
+def main():
+    spectra = run_sims()
+    fs = np.sum(spectra, axis=0)
+    fs = moments.Spectrum(fs, pop_ids=_pop_ids)
+    fs.tofile("data/data.fs")
+
+    bs_spectra = bootstrap_spectra(spectra)
+    with gzip.open("data/bootstrapped_data.pkl.gz", "wb+") as fout:
+        pickle.dump(bs_spectra, fout)
+
+if __name__ == "__main__":
+    main()

### 2025-03-18 example2/models/MSL/plot_model.py
finished setting up example2 examples

@@ -21,7 +21,8 @@ def main():
     data = moments.Spectrum.from_file(data_file)
 
     # plot size history
-    demesdraw.size_history(graph)
+    ax = plt.subplot(111)
+    demesdraw.size_history(graph, ax=ax)
     plt.savefig("MSL_model.misid_fit.size_history.png")
     plt.close()
 

### 2025-03-18 example2/models/MSL_GBR/plot_model.py

@@ -21,7 +21,8 @@ def main():
     data = moments.Spectrum.from_file(data_file)
 
     # plot size history
-    demesdraw.size_history(graph, log_time=True)
+    ax = plt.subplot(111)
+    demesdraw.size_history(graph, ax=ax)
     plt.savefig("MSL_GBR_model.misid_fit.size_history.png")
     plt.close()
 

### 2025-03-18 example2/models/MSL_GBR_Vindija_round1/fit_model.py

@@ -8,7 +8,7 @@ def main():
     graph_file = "MSL_GBR_Vindija_model.yaml"
     options_file = "options_MSL_GBR_Vindija.yaml"
     data_file = "../../spectra/MSL_GBR_Vindija"
-    output = "fits/MSL_GBR_Vindija_model.misid_fit.yaml"
+    output = "MSL_GBR_Vindija_model.misid_fit.yaml"
     u = 1.5e-8
     L = 960914001
     misid_guess = 0.03

### 2025-03-18 example2/models/MSL_GBR_Vindija_round2/compute_uncerts.py

@@ -1,0 +1,42 @@
+## Compute uncertainties using the FIM method
+
+import demes
+import moments 
+
+
+def main():
+    # define paths and parameters
+    graph_file = "MSL_GBR_Vindija_round2.misid_fit.yaml"
+    options_file = "options_MSL_GBR_Vindija_round2.yaml"
+    data_file = "../../spectra/MSL_GBR_Vindija"
+    u = 1.5e-8
+    L = 960914001
+    p_misid = 0.0223
+
+    data = moments.Spectrum.from_file(data_file)
+
+    # load parameter names and best-fit values using moments.Demes functions
+    builder = demes.load(graph_file).asdict()
+    options = moments.Demes.Inference._get_params_dict(options_file)
+    param_names, fit_params, _, __ = \
+        moments.Demes.Inference._set_up_params_and_bounds(options, builder)
+
+    std_errs = moments.Demes.Inference.uncerts(
+        graph_file,
+        options_file,
+        data,
+        uL=u * L,
+        fit_ancestral_misid=True,
+        misid_fit=p_misid,
+    )
+
+    # print results
+    print(r"95% confidence intervals:")
+    print("param\t2.5%\t97.5%")
+    for name, val, err in zip(param_names, fit_params, std_errs):
+        print(f"{name}\t{val - 1.96 * err:.3}\t{val + 1.96 * err:.3}")
+
+    return 
+
+
+main() 

### 2025-03-18 example2/models/MSL_GBR_Vindija_round2/fit_model.py

@@ -1,31 +1,34 @@
-## Fit a model of MSL, GBR, and Vindija populations.
+## Refit the MSL, GBR, Vindija model to fine-tune parameters
 
 import moments 
-import sys
 
 
 def main():
     # define paths and parameters
-    tag = sys.argv[1]
-    graph_file = "MSL_GBR_Vindija_model3.yaml"
-    options_file = "options_MSL_GBR_Vindija3.yaml"
+    graph_file = "MSL_GBR_Vindija_round2.yaml"
+    options_file = "options_MSL_GBR_Vindija_round2.yaml"
+    data_file = "../../spectra/MSL_GBR_Vindija"
+    output = "MSL_GBR_Vindija_round2.misid_fit.yaml"
     u = 1.5e-8
     L = 960914001
-    data = moments.Spectrum.from_file("../../data/data_MSL_GBR_Vindija")
-    output = f"MSL_GBR_Vindija_round3_misid_best_fit_{tag}.yaml"
+    misid_guess = 0.024
+
+    data = moments.Spectrum.from_file(data_file)
+
     param_names, misid_popt, misid_ll = moments.Demes.Inference.optimize(
         graph_file, 
         options_file, 
         data, 
         maxiter=10000,
         fit_ancestral_misid=True, 
-        misid_guess=0.03,
+        misid_guess=misid_guess,
         uL=u * L, 
         verbose=1,
-        output=output, 
-        perturb=0.25,
-        method="powell"
+        output=output,
+        method="lbfgsb"
     )
+
+    # print results
     print("Best-fit log-likelihood:", -misid_ll)
     print("Best-fit parameters:")
     for name, popt in zip(param_names, misid_popt):

### 2025-03-18 example2/models/MSL_GBR_Vindija_round2/fit_model.sh

@@ -1,2 +1,0 @@
-#!/bin/bash
-python fit_model_remote.py $1

### 2025-03-18 example2/models/MSL_GBR_Vindija_round2/fit_model_remote.py

@@ -1,37 +1,0 @@
-## Fit a model of MSL, GBR, and Vindija populations.
-
-import moments 
-import sys
-
-
-def main():
-    # define paths and parameters
-    tag = sys.argv[1]
-    graph_file = "MSL_GBR_Vindija_model3.yaml"
-    options_file = "options_MSL_GBR_Vindija3.yaml"
-    u = 1.5e-8
-    L = 960914001
-    data = moments.Spectrum.from_file("data_MSL_GBR_Vindija")
-    output = f"MSL_GBR_Vindija_round3_misid_best_fit_{tag}.yaml"
-    param_names, misid_popt, misid_ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=0.03,
-        uL=u * L, 
-        verbose=1,
-        output=output, 
-        perturb=0.25,
-        method="powell"
-    )
-    print("Best-fit log-likelihood:", -misid_ll)
-    print("Best-fit parameters:")
-    for name, popt in zip(param_names, misid_popt):
-        print(f"{name}\t{popt:.3}")
-
-    return 
-
-
-main() 

### 2025-03-18 example2/models/MSL_GBR_Vindija_round2/plot_model.py

@@ -8,29 +8,32 @@ import moments
 
 def main():
 
-    graph_file = "MSL_GBR_Vindija_model.misid_fit.yaml"
+    graph_file = "MSL_GBR_Vindija_round2.misid_fit.yaml"
     data_file = "../../spectra/MSL_GBR_Vindija"
     u = 1.5e-8
     L = 960914001
-    p_misid = 0.0250059
+    p_misid = 0.0218
 
     graph = demes.load(graph_file)
     data = moments.Spectrum.from_file(data_file)
     model = moments.Demes.SFS(
         graph, 
-        samples={"MSL":20, "GBR":20, "Vindija": 2}, 
+        samples={"MSL": 20, "GBR": 20, "Vindija": 2}, 
         u=u,
         L=L
     )
     model = moments.Misc.flip_ancestral_misid(model, p_misid)
 
-    # plot demography in linear and log time
-    demesdraw.tubes(graph, log_time=True)
-    plt.savefig("MSL_GBR_Vindija_model.misid_fit.tubes_logtime.png")
-    plt.close()
+    print(moments.Inference.ll(model, data))
 
-    demesdraw.tubes(graph)
-    plt.savefig("MSL_GBR_Vindija_model.misid_fit.tubes.png")
+    # plot demography in linear and log time
+    ax = plt.subplot(111)
+    demesdraw.tubes(graph, log_time=True, ax=ax)
+    plt.savefig("MSL_GBR_Vindija_round2.misid_fit.tubes_logtime.png")
+    plt.close()
+    ax = plt.subplot(111)
+    demesdraw.tubes(graph, ax=ax)
+    plt.savefig("MSL_GBR_Vindija_round2.misid_fit.tubes.png")
     plt.close()
 
     # plot marginal fit for GBR
@@ -38,15 +41,22 @@ def main():
         model.marginalize([0, 2]), 
         data.marginalize([0, 2]),
         residual="linear", 
-        out="MSL_GBR_Vindija_model.misid_fit.comp_1d_GBR.png"
-    )   
+        out="MSL_GBR_Vindija_round2.misid_fit.comp_1d_GBR.png"
+    )
+    # and for MSL
+    moments.Plotting.plot_1d_comp_Poisson(
+        model.marginalize([1, 2]), 
+        data.marginalize([1, 2]),
+        residual="linear", 
+        out="MSL_GBR_Vindija_round2.misid_fit.comp_1d_MSL.png"
+    )
 
[…]

### 2025-03-20 example2/data/tools/get_ancestral_states.py
added readmes and snakefiles in data/

@@ -10,6 +10,8 @@ import argparse
 import gzip 
 import numpy as np
 
+from lib import *
+
 
 def get_args():
 
@@ -25,34 +27,6 @@ def get_args():
     )
 
     return parser.parse_args()
-
-
-def load_fa_array(file):
-    """
-    Load the characters in a .fa file and return them as a numpy array.
-    """
-    lines = []
-    line0 = None
-    with open(file, "r") as fin:
-        for line in fin:
-            if ">" in line:
-                line0 = line
-                continue
-            lines.append(line.rstrip("\n"))
-    array = np.array([c for c in "".join(lines)])
-
-    return line0, array
-
-
-def write_tab_file(file, positions, states, chromnum):
-
-    openfunc = gzip.open if file.endswith(".gz") else open 
-    with openfunc(file, "wb") as fout:
-        fout.write("#CHROM\tPOS\tSTATE\n".encode())
-        for pos, state in zip(positions, states):
-            lineb = f"{chromnum}\t{pos}\t{state}\n".encode()
-            fout.write(lineb)
-    return
 
 
 def main():

### 2025-03-20 example2/data/tools/get_vcf_coverage.py

@@ -7,6 +7,8 @@ import argparse
 import gzip
 import numpy as np 
 
+from lib import *
+
 
 def get_args():
 
@@ -18,18 +20,6 @@ def get_args():
         "-o", "--out_file", required=True
     )
     return parser.parse_args()
-
-
-def write_bedfile(file, regions, chromnum):
-    """
-    write a .bed file (supports .gz) from an array of mask regions.
-    """
-    openfunc = gzip.open if file.endswith(".gz") else open 
-    with openfunc(file, "wb") as fout:
-        for start, end in regions:
-            lineb = f"{chromnum}\t{start}\t{end}\n".encode()
-            fout.write(lineb)
-    return
 
 
 def main():

### 2025-03-20 example2/data/tools/lib.py

@@ -1,4 +1,4 @@
-## utilities used by other scripts
+## utilities used by scripts
 
 import gzip
 import numpy as np
@@ -33,7 +33,10 @@ def boolmask_to_regions(boolmask):
 
 
 def read_bedfile(file):
-
+    """
+    Read a .bed file and return an array of regions and the chromosome number,
+    represented as a string.
+    """
     openfunc = gzip.open if file.endswith(".gz") else open 
     regions = []
     with openfunc(file, "rb") as fin:
@@ -118,3 +121,33 @@ def subtract_regions(regions, subtrahend):
     ret = boolmask_to_regions(outmask)
 
     return ret
+
+
+def load_fa_array(file):
+    """
+    Load the characters in a .fa file and return them as a numpy array.
+    """
+    lines = []
+    line0 = None
+    with open(file, "r") as fin:
+        for line in fin:
+            if ">" in line:
+                line0 = line
+                continue
+            lines.append(line.rstrip("\n"))
+    array = np.array([c for c in "".join(lines)])
+
+    return line0, array
+
+
+def write_tab_file(file, positions, states, chromnum):
+    """
+    Write ancestral states to a .tab file with columns CHROM, POS, STATE.
+    """
+    openfunc = gzip.open if file.endswith(".gz") else open 
+    with openfunc(file, "wb") as fout:
+        fout.write("#CHROM\tPOS\tSTATE\n".encode())
+        for pos, state in zip(positions, states):
+            lineb = f"{chromnum}\t{pos}\t{state}\n".encode()
+            fout.write(lineb)
+    return

### 2025-03-20 example2/data/tools/parse_sfs.py

@@ -1,10 +1,11 @@
-## parse the SFS from a .vcf file with ancestral state annotations, then save it
+## Parse the SFS from a .vcf file with ancestral state annotations, then save it
+## Uses a function adapted from the `from_data_dict` method of the Spectrum
+## class in moments.Spectrum_mod.
 
 import argparse
 from collections import defaultdict
 import moments 
 import numpy as np
-import pickle 
 
 
 def get_args():
@@ -25,7 +26,8 @@ def get_args():
 
 def compute_sfs(data_dict, pop_ids, sample_sizes, mask_corners=True):
     """
-    copied from moments
+    Compute the SFS from a data dictionary. Details on this format can be found
+    in the moments.Misc module.
     """
     Npops = len(pop_ids)
     fs = np.zeros(np.asarray(sample_sizes) + 1)

### 2025-03-20 example2/data/tools/sum_sfs.py

@@ -1,10 +1,10 @@
-## add together spectra from several saved .sfs files and write the sum
+## add together spectra from several saved .sfs files and write the sum. Also
+## marginalizes to the populations MSL, CHS, GBR, Vindija
 
 import argparse
 from collections import defaultdict
 import moments 
 import numpy as np
-import pickle 
 
 
 def get_args():

### 2025-03-25 chapter/paper.tex
add some text, a figure

--- 2025-03-18 chapter/paper.tex
@@
 % need to have the pythonhighlight.sty in directory
-\usepackage{pythonhighlight}
+%\usepackage{pythonhighlight}
+
+% use minted for python and YAML code
+\usepackage{minted}
 
@@
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
+\usepackage{xspace}
+\newcommand{\moments}{\texttt{moments}\xspace}
+\newcommand{\demes}{\texttt{demes}\xspace}
+\newcommand{\msprime}{\texttt{msprime}\xspace}
+\newcommand{\momi}{\texttt{momi}\xspace}
+\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
+\newcommand{\tskit}{\texttt{tskit}\xspace}
 
@@
 
-\title{Demographic inference using the SFS with Moments and Demes}
+\title{Demographic inference using the SFS with \moments and \demes}
 \author[1]{Nick Collier}
@@
 
-\abstract{
+\begin{abstract}
     Placeholder
-}
+\end{abstract}
 
-\section{Introduction}
+\section*{Introduction}
 
-What moments is, what type of data it uses, general framework for inference
-\begin{enumerate}
-    \item The SFS -- how its shape is sensitive to population,
-        selection and genome biological parameters
-    \item Demographic inference from the SFS
-    \item Extensions: inference of selection, mutation rates, etc
-    \item Extensions: Multi-allele and multi-locus models
-\end{enumerate}
+The genetic composition of a sample of individuals is shaped by their genome
+biology and evolutionary history. Variation resulting from this history can be
+fully represented by the ancestral relationships among samples at each locus in
+the genome and how those gene-genealogical relationships change along a
+chromosome due to recombination (that is, information stored in the Ancestral
+Recombination Graph \citep{nielsen2025inference}). However, the ARG can be
+large and unwieldy, and methods for reconstructing history directly from the
+ARG, while showing promise \citep[e.g.,][]{yc2022evaluation, fan2023likelihood,
+brandt2024promise}, are in their infancy and so far limited in application and
+scalability. Instead, evolutionary inference using informative summaries of
+genetic variation remains a tractable and powerful alternative for learning
+parameters of population history, natural selection and genome biology.
 
-\section{Interface with Demes}
+One such summary that has seen wide use is the site frequency spectrum (SFS),
+which stores the counts (observed or expected) of alleles carried by a given
+number of genomes in a set of samples. Like any summary of the data, the SFS
+discards some information stored in the ARG -- in this case, loci are treated
+independently so that haplotypic information is lost. Even so, the relative
+densities of allele frequencies and the overall scale of the SFS are both
+sensitive to demographic and non-demographic processes.
+
+This chapter focuses on multi-population demographic inference from the SFS
+using \moments \citep{jouganous2017inferring}.  Past population processes,
+including changes in population size, splits, gene flow and population
+structure, impact the SFS in predictable ways. A population size expansion (or
+contraction) leads to a relative excess (or deficit) of low-frequency variants,
+for example, and commonly used statistics measuring population divergence, such
+as $F_{ST}$, are themselves summaries of the SFS.  Thus, the development of
+methods to learn demographic history from the SFS came soon after the
+publication of whole-genome sequencing from multiple individuals
+\citep{marth2004allele, williamson2005simultaneous}. More recent methodological
+advances allow for increased sample sizes and numbers of populations, providing
+a rich ecosystem of software for SFS-based demographic inference
+\citep{gutenkunst2009inferring, excoffier2011fastsimcoal,
+gravel2011demographic, jouganous2017inferring, ragsdale2018genomic,
+kamm2020efficiently, dilber2024faster}. We return to this in the final section:
+Considerations and Caveats.
+
+Other evolutionary mechanisms are known to affect allele frequency dynamics and
+the SFS. This presents a challenge for demographic inference, as the observed
+SFS may be distorted by non-demographic processes. However, this also presents
+an opportunity to learn about different evolutionary processes, if demographic
+history can be controlled for. The SFS has been used to infer the distribution
+of fitness effects of new mutations, primarily for nonsynonymous variation in
+coding regions \citep[e.g.,][]{eyre2006distribution, boyko2008assessing,
+kim2017inference}; to scan for historical selective sweeps
+\citep[e.g.,][]{kim2002detecting, nielsen2005genomic}; and to understand how
+selection on quantitative traits impacts the genetic architecture underlying
+those traits \citep[eg.,][]{patel2024conditional, ragsdale2024archaic}. The SFS
+has also been used to infer relative mutation rates and how they have changed
+over time \citep{dewitt2021nonparametric}. Each of these analyses typically
+requires partitioning the data in some way -- by functional annotation,
+mutation context, local recombination rate, etc -- so that comparisons can be
+made across SFS from different classes of mutations.
+
+\subsection*{Interfacing \moments with \demes}
+
+Specifying demographic models requires defining populations (or ``demes'') and
+their relationships via splits and gene flow. \demes provides a standardized
+and accessible format for defining demographic models \citep{gower2022demes},
+and has been adopted by widely used software for population genetics simulation
+and inference (including \msprime \citep{baumdicker2022efficient}, \momi
+\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, \texttt{GADMA}
+\citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
+code across software, and ease of model implementation and reduction of errors.
+
+Demographic models are specified in YAML format (see \citet{gower2022demes} and
+the associated documentation). The two-population split-with-migration model
+depicted in Figure~\ref{fig:im}A is specified as below:
+\begin{minted}[gobble=4, frame=single,]{yaml}
+    description: split-with-migration model, loosely based on example 2
+    generation_time: 29
+    time_units: years
+    demes:
+    - name: ancestral
+      epochs:
+      - {start_size: 15000, end_time: 300000}
+      - {start_size: 25000, end_time: 75000}
+    - name: popA
+      ancestors: [ancestral]
+      epochs:
+      - {start_size: 30000}
+    - name: popB
+      ancestors: [ancestral]
+      epochs:
+      - {start_size: 1200, end_size: 14000}
+    migrations:
+    - demes: [popA, popB]
+      rate: 5e-5
+\end{minted}
+
+In \moments, a \demes-specified demographic model can be directly used to
+compute either the SFS or multi-population LD statistics
+\citep{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
+to load the demographic model using \demes and specify the number of haploid
+samples to draw and from which populations.
+\begin{minted}[gobble=4, frame=single]{python}
+    import demes, moments
+    g = demes.load('model.yaml')
+    samples = {'popA': 60, 'popB': 60}
+    fs = moments.Demes.SFS(g, samples=samples)
+\end{minted}
+Here, we have not specified a mutation rate. The default behavior sets $u=1$,
+so that $\theta = 4N_eu=4N_e$, where $N_e$ is the ancestral population's
+initial size. With a known mutation rate, the SFS will be properly scaled by
+passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
+samples=samples, u=u)}. This is equivalent to scaling the output SFS above by
+multiplying by \texttt{u}, as \texttt{u*fs}.
+
+\section*{Examples: inference in \moments using \demes}
+
+Below, we show two examples of multi-population demographic inference using the
+joint SFS. The first example simulates data with known demographic parameters,
+which we try to reinfer using the both the original and misspecified simpler
+models. In the second example, we infer a relatively simple human-Neanderthal
+history using data from two human populations and a high-coverage ancient
+sample from the Neanderthal lineage. For both of these examples, we briefly
+describe data processing and inference setup, highlighting the high-level
+components of the analyses. For detailed information for each example,
+including managing data, specifying models, performing inference, computing
+confidence intervals and visualizing fits, we refer readers to the accompanying
+GitHub repository
+(\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
+is maintained with up-to-date versioning.
+
+\begin{itemize}
+    \item to run inference, need to specify parameters to be fit
+    \item the parameters/options file points to values in the demographic model
+        to be fit, along with bounds on those parameters and any constraints
+        among parameters
+    \item this options file is also a YAML -- see snippets of examples below, and
+        full details are given in the above URL
+\end{itemize}
+
+\subsection*{Inferring parameters in a simulated split-with-migration model}
+
+Using the split-with-migration model defined above, we simulated data for 30
+diploid individuals from both populations using \msprime
+\citep{baumdicker2022efficient}. These simulations consided of 500 regions,
+each of length 1 Mb, with per-base recombination and mutation rates of
+$10^{-8}$. The joint SFS was computed using \tskit's
+\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
+region and summing across regions to find the total SFS across 500 Mb of data
+(Figure~\ref{fig:im}B--D).
+
+\begin{figure}[t!]
+    \includegraphics{../example1/fig1.pdf}
+    \caption{Caption placeholder}
+    \label{fig:im}
+\end{figure}
+
+\subsection*{Inferring human-Neanderthal demographic parameters}
+
+\begin{figure}[t!]
+    %\includegraphics{../example2/fig2.pdf}
+    \caption{Caption placeholder}
+    \label{fig:humans}
+\end{figure}
+
+\section*{Considerations and caveats}\label{sec:conclusions}
 
 \begin{enumerate}
-    \item Manipulating and plotting Demes-specified demographic models
-    \item Computing statistics given a Demes model
-    \item Inferring history on a demes model, including specifying parameters to be fit
+    \item Include here general ideas about the strengths and weaknesses of
+        various approaches in population genetic inference, including when
+        using the SFS.
+    \item Challenges in finding local/global optima.
+    \item Challenges in exploring parameter space.
+    \item Background selection [Ewing, Johri]
+    \item Gene dense vs gene sparse genomic architectures among species.
 \end{enumerate}
-
-\section{Examples}
-
-Two examples, one using simulated data with known parameters, which we try to
-reinfer. And another with actual data (humans + neanderthal). We briefly
-describe these, highlighting the high-level components of the fits. For
-detailed information for each, including managing data, specifying models,
-performing inference, and visualizing results, we refer readers to the GitHub
-(\url{}), which is maintained with up-to-date versioning.
-
-\subsection{Inferring parameters in a simulated isolation-with-migration model}
-
-A python snippet:
-\begin{python}
-dbg = msprime.DemographyDebugger(
-  population_configurations=population_configurations,
-  migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
-dbg.print_history()
-ts = msprime.simulate(
-  population_configurations=population_configurations,
-  migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
-\end{python}
-or
-\begin{python}
-dbg = demography.debug()
-dbg.print_history()
-ts = msprime.simulate(demography=demography)
-\end{python}
-
-This was in discussion of \citet{kelleher2016efficient}.
-
-\subsection{Inferring human-Neanderthal demographic parameters}
-
-\begin{python}
-def myFunc():
-    pass
-\end{python}
-
-\section{Considerations and caveats}
-
-Incluce here general ideas about the strengths and weaknesses of various approaches
-in population genetic inference, including when using the SFS.
-
-Challenges in finding local/global optima. Challenges in exploring parameter space.
 
@@
 
-\clearpage
-
-\appendix
-\renewcommand{\thesection}{A\arabic{section}}
-\renewcommand{\theequation}{A\arabic{equation}}
-\renewcommand{\thefigure}{A\arabic{figure}}
-\renewcommand{\thetable}{A\arabic{table}}
-\setcounter{figure}{0}
-\setcounter{equation}{0}
-\setcounter{table}{0}
-
-%\section{Appendix} \label{appendix}
 \end{document}

### 2025-03-25 example1/plot_fig1.py

@@ -1,0 +1,132 @@
+import demes, demesdraw
+import matplotlib, matplotlib.pylab as plt
+import numpy as np
+import moments
+
+# set font sizes
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=5)
+matplotlib.rc("ytick", labelsize=5)
+matplotlib.rc("axes", labelsize=6)
+matplotlib.rc("axes", titlesize=6)
+matplotlib.rc("legend", fontsize=5)
+
+
+def plot_demes_model(g, ax, inf_ratio=0.2, top_lim=None, ylabel=None, title=None):
+    demesdraw.tubes(g, ax=ax, labels="xticks", inf_ratio=inf_ratio)
+    if top_lim is not None:
+        ax.set_ylim(top=top_lim)
+    if ylabel is not None:
+        ax.set_ylabel(ylabel)
+    if title is not None:
+        ax.set_title(title)
+
+def plot_1d_sfs(data, ax):
+    pass
+
+def plot_2d_sfs(data, ax):
+    pass
+
+def plot_2d_resid(data, model, ax):
+    pass
+
+
+grid = (4, 9)
+
+g = demes.load("models_and_fits/model.yaml")
+g1 = demes.load("models_and_fits/model.misspec.out.yaml")
+g2 = demes.load("models_and_fits/model.misspecB.out.yaml")
+g3 = demes.load("models_and_fits/model.out.yaml")
+
+data = moments.Spectrum.from_file("data/data.fs")
+
+model1 = moments.Demes.SFS(g1, samples={"popA": 60, "popB": 60}, u=500*1e-8*1e6)
+resid1 = moments.Inference.Anscombe_Poisson_residual(model1, data)
+
+model2 = moments.Demes.SFS(g2, samples={"popA": 60, "popB": 60}, u=500*1e-8*1e6)
+resid2 = moments.Inference.Anscombe_Poisson_residual(model2, data)
+
+model3 = moments.Demes.SFS(g3, samples={"popA": 60, "popB": 60}, u=500*1e-8*1e6)
+resid3 = moments.Inference.Anscombe_Poisson_residual(model3, data)
+
+fig = plt.figure(1, figsize=(6.5, 5), dpi=300)
+fig.clf()
+
[…]

### 2025-03-26 chapter/paper.tex
it's not good writing, but it's writing

@@
 
-\begin{itemize}
-    \item to run inference, need to specify parameters to be fit
-    \item the parameters/options file points to values in the demographic model
-        to be fit, along with bounds on those parameters and any constraints
-        among parameters
-    \item this options file is also a YAML -- see snippets of examples below, and
-        full details are given in the above URL
-\end{itemize}
-
-\subsection*{Inferring parameters in a simulated split-with-migration model}
-
-Using the split-with-migration model defined above, we simulated data for 30
-diploid individuals from both populations using \msprime
-\citep{baumdicker2022efficient}. These simulations consided of 500 regions,
-each of length 1 Mb, with per-base recombination and mutation rates of
-$10^{-8}$. The joint SFS was computed using \tskit's
-\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
-region and summing across regions to find the total SFS across 500 Mb of data
-(Figure~\ref{fig:im}B--D).
-
 \begin{figure}[t!]
@@
 \end{figure}
+
+In order to optimize parameters in a \demes model, we need a way to specify
+which parameters should be fit. For this, we use a separate YAML-formatted file
+to define \texttt{parameters}, each of which points to the value(s) in the
+demographic model and sets lower and upper bounds. This file also specifies any
+relative \texttt{constraints} between pairs of parameters, to ensure the
+optimization routine only explores valid model space. Below, we have included
+a truncated parameters file -- the full example is available on GitHub.
+\begin{minted}[gobble=4, frame=single]{YAML}
+    parameters:
+    - name: T0
+      values:
+      - demes:
+          ancestral:
+            epochs:
+              0: end_time
+      lower_bound: 0
+      upper_bound: 2e5
+    - name: T1
+      values:
+      - demes:
+          ancestral:
+            epochs:
+              1: end_time
+      lower_bound: 0
+      upper_bound: 1e6
+    - name: Ne
+      values:
+      - demes:
+          ancestral:
+            epochs:
+              0: start_size
+      lower_bound: 1e2
+      upper_bound: 1e6
+    ...
+    constraints:
+    - params: [T0, T1]
+      constraint: greater_than
+\end{minted}
+
+Taking the demographic model, parameters file, and observed SFS together, we
+can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
+function.  The initial parameter guesses are given by the input demographic
+model, which may be perturbed using a keyword argument. Any value in the
+demographic model that is not specified in the parameters file will remain
+unchanged, and is therefore treated as a fixed parameter. Here, we also assume
+we have an estimate for the total mutation rate, $U$.
+\begin{minted}[gobble=4, frame=single]{YAML}
+    import moments
+
+    data = moments.Spectrum.from_file('data.fs')
+    U = 1e-8 * 5e8  # per-base mutation rate times the total length
+
+    model_file = 'model.yaml'
+    params_file = 'params.yaml'
+    output = 'model.fit.yaml'
+
+    ret = moments.Demes.Inference.optimize(
+        model_file, params_file, data,
+        perturb=1, uL=U, output=output
+    )
+\end{minted}
+
+Confidence intervals may also be computed, once a (locally) optimal model has
+been found.
+
+\subsection*{Inferring parameters in a simulated split-with-migration model}
+
+Using the split-with-migration model defined above (illustrated in
+Figure~\ref{fig:im}A), we simulated data for 30 diploid individuals from both
+populations using \msprime \citep{baumdicker2022efficient}. These simulations
+consided of 500 regions, each of length 1 Mb, with per-base recombination and
+mutation rates of $10^{-8}$. The joint SFS was computed using \tskit's
+\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
+region and summing across regions to find the total SFS across 500 Mb of data
+(Figure~\ref{fig:im}B--D).
+
+We fit three demographic models to the simulated data. In addition to
+reinferring the parameters from the simulated model, we fit two simpler models
+to the same data. These had fewer parameters, both omitting the size change
+deeper in time in the ancestral population. This is meant to crudely mimic the
+scenario that we often face, in which the true history is more complicated than
+the parameterized model. It also highlights biases in inferences that can arise
+when features of the true history are not included.
+
+When fitting the two misspecified models that do not allow for size changes in
+the ancestral population, this split time is inferred to be substantially
+deeper in the past than the true split time. This effect, as demonstrated by
+\citet{momigliano2020biases}, is due to the decreased coalescence rate within
+the ancestral population between the time of the expansion and divergence of
+the descendent populations. These two misspecified models
+(Figure~\ref{fig:im}H,I) also provide a worse fit to the data, as seen by the
+lower log-likelihoods and the large residuals between model predictions and
+data (Figure~\ref{fig:im}E-G).
 

### 2025-03-27 chapter/paper.tex
wrote a little about example 2

@@
 
+We obtained modern human genome sequences from a recent resequencing of the 1000 Genomes Project cohort \citep{byrska2022high}, and a high-coverage sequence of the Vindija Neandertal from \cite{prufer2017high}.
+From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR (British from England and Scotland) sequences as our modern human samples.  
+After filtering out sites that lay outside the 1000 Genomes `strict' mask, were not genotyped in all four existing high-coverage archaic genome sequences, lacked high-confidence ancestral state assignments, or fell within 10 kb of exonic or promoter regions, we were left with approximately 960 Mb of well-characterized and putatively neutrally evolving sequence.
+We estimated the genome-wide SFS using built-in \moments parsing functions.
+Details concerning data processing (including our liftover of the Vindija sequence to a more recent genome build) can be found in the linked GitHub repository.
+
+The model that we wished to fit is relatively simple but still has many parameters (11 in the example we present), so we constructed it iteratively by first dealing with simpler one- and two-population graphs.
+We began by fitting one-population models with ancestral and recent size changes for MSL and GBR.
+We attributed to GBR a sharp contraction ~60 kya followed by exponential growth to represent the out-of-Africa bottleneck and subsequent rapid expansion.
+Then we stitched these one-population models together and refitted to get a simple two-population out-of-Africa model. 
+Following this, we added a Neandertal branch and fit onlt the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixuture proportion. 
+We also fitted the divergence time between the sampled Vindija Neandertal and the lineage which admixed with modern humans, which we observed to be poorly constrained.
+In a second round of optimization, we fixed the aforesaid divergence time and refit all other parameters to refine the model.
+
 \begin{figure}[t!]

### 2025-03-27 example2/models/MSL_GBR_Vindija_round3/compute_uncerts.py

@@ -6,8 +6,8 @@ import moments
 
 def main():
     # define paths and parameters
-    graph_file = "MSL_GBR_Vindija_round2.misid_fit.yaml"
-    options_file = "options_MSL_GBR_Vindija_round2.yaml"
+    graph_file = "MSL_GBR_Vindija_round3.misid_fit.yaml"
+    options_file = "options_MSL_GBR_Vindija_round3.yaml"
     data_file = "../../spectra/MSL_GBR_Vindija"
     u = 1.5e-8
     L = 960914001

### 2025-03-27 example2/models/MSL_GBR_Vindija_round3/fit_model.py

@@ -1,0 +1,40 @@
+## Refit the MSL, GBR, Vindija model to fine-tune parameters
+
+import moments 
+
+
+def main():
+    # define paths and parameters
+    graph_file = "MSL_GBR_Vindija_round3.yaml"
+    options_file = "options_MSL_GBR_Vindija_round3.yaml"
+    data_file = "../../spectra/MSL_GBR_Vindija"
+    output = "MSL_GBR_Vindija_round3.misid_fit.yaml"
+    u = 1.5e-8
+    L = 960914001
+    misid_guess = 0.024
+
+    data = moments.Spectrum.from_file(data_file)
+
+    param_names, misid_popt, misid_ll = moments.Demes.Inference.optimize(
+        graph_file, 
+        options_file, 
+        data, 
+        maxiter=10000,
+        fit_ancestral_misid=True, 
+        misid_guess=misid_guess,
+        uL=u * L, 
+        verbose=1,
+        output=output,
+        method="lbfgsb"
+    )
+
+    # print results
+    print("Best-fit log-likelihood:", -misid_ll)
+    print("Best-fit parameters:")
+    for name, popt in zip(param_names, misid_popt):
+        print(f"{name}\t{popt:.3}")
+
+    return 
+
+
+main() 

### 2025-03-27 example2/models/MSL_GBR_Vindija_round3/plot_model.py

@@ -1,0 +1,67 @@
+## Make model/data comparison plots for the model fit by ./fit_model.py
+
+import demes
+import demesdraw 
+import matplotlib.pyplot as plt
+import moments
+
+
+def main():
+
+    graph_file = "MSL_GBR_Vindija_round3.misid_fit.yaml"
+    data_file = "../../spectra/MSL_GBR_Vindija"
+    u = 1.5e-8
+    L = 960914001
+    p_misid = 0.0223
+
+    graph = demes.load(graph_file)
+    data = moments.Spectrum.from_file(data_file)
+    model = moments.Demes.SFS(
+        graph, 
+        samples={"MSL": 20, "GBR": 20, "Vindija": 2}, 
+        u=u,
+        L=L
+    )
+    model = moments.Misc.flip_ancestral_misid(model, p_misid)
+
+    print(moments.Inference.ll(model, data))
+
+    # plot demography in linear and log time
+    ax = plt.subplot(111)
+    demesdraw.tubes(graph, log_time=True, ax=ax)
+    plt.savefig("MSL_GBR_Vindija_round3.misid_fit.tubes_logtime.png")
+    plt.close()
+    ax = plt.subplot(111)
+    demesdraw.tubes(graph, ax=ax)
+    plt.savefig("MSL_GBR_Vindija_round3.misid_fit.tubes.png")
+    plt.close()
+
+    # plot marginal fit for GBR
+    moments.Plotting.plot_1d_comp_Poisson(
+        model.marginalize([0, 2]), 
+        data.marginalize([0, 2]),
+        residual="linear", 
+        out="MSL_GBR_Vindija_round3.misid_fit.comp_1d_GBR.png"
+    )
+    # and for MSL
+    moments.Plotting.plot_1d_comp_Poisson(
+        model.marginalize([1, 2]), 
+        data.marginalize([1, 2]),
+        residual="linear", 
+        out="MSL_GBR_Vindija_round3.misid_fit.comp_1d_MSL.png"
+    )
+
+    # plot 3d comparison
+    moments.Plotting.plot_3d_comp_Poisson(
+        model, 
+        data, 
+        residual="linear", 
+        out="MSL_GBR_Vindija_round3.misid_fit.comp_3d.png"
+    )
+
+    return
+
+
+if __name__ == "__main__":
+    main()
+

### 2025-03-27 chapter/paper.tex
added figure 2

--- 2025-03-26 chapter/paper.tex
@@
 \newcommand{\demes}{\texttt{demes}\xspace}
+\newcommand{\demesdraw}{\texttt{demesdraw}\xspace}
 \newcommand{\msprime}{\texttt{msprime}\xspace}
@@
 
+We obtained modern human genome sequences from a recent resequencing of the 1000 Genomes Project cohort \citep{byrska2022high}, and a high-coverage sequence of the Vindija Neandertal from \cite{prufer2017high}.
+From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR (British from England and Scotland) sequences as our modern human samples.  
+After filtering out sites that lay outside the 1000 Genomes `strict' mask, that were not genotyped in all four existing high-coverage archaic genome sequences, lacked high-confidence ancestral state assignments, or fell within 10 kb of exonic or promoter regions, we were left with approximately 960 Mb of well-characterized and putatively neutrally-evolving sequence.
+We estimated the genome-wide SFS using built-in \moments parsing functions.
+Details concerning data processing (including our liftover of the Vindija sequence to a more recent genome build) can be found in the linked GitHub repository.
+
+The model that we wished to fit is relatively simple but still has many parameters (11 in the example we present), so we constructed it in stages by first fitting simpler one- and two-population graphs.
+Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per generation and 29 years as the average generation time.
+We began by fitting one-population models with ancestral and recent size changes for MSL and GBR.
+For the MSL model, we allowed three epochs with constant effective size in each epoch.
+We attributed to GBR a sharp contraction $\sim$60 kya followed by exponential growth to represent the out-of-Africa bottleneck and subsequent rapid expansion.
+YAML files representing the models and parameters in this section can be found in the linked GitHub repository.
+
+We next stitched these one-population models together and refitted the resulting graph to get a simple two-population out-of-Africa model, with continuous symmetric migration between MSL and GBR. 
+Following this, we added a Neandertal branch and fit only the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixture proportion. 
+We also fitted the divergence time between the sampled Vindija Neandertal and the lineage which admixed with modern humans, which we observed to be poorly constrained, likely in part due to the very small size of our Vindija sample.
+In a second round of optimization for the three-population model, we fixed the aforesaid divergence time to a value supported by literature (90 kya, \cite{prufer2017high}) and refit all other parameters to refine the model.
+
+After this refitting step, we were left with parameters in good agreement with prior authors, with a Neandertal-modern human divergence time of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 percent.
+We observe some large residuals in our best-fit model, suggesting ...
+
 \begin{figure}[t!]
-    %\includegraphics{../example2/fig2.pdf}
-    \caption{Caption placeholder}
+    \includegraphics{../example2/fig2.pdf}
+    \caption{
+      A plot of the best-fit model created using \demesdraw is displayed in (A).
+      Marginal SFS fits between the model and empirical data are shown for MSL (B) and GBR (C).
+      Marginal two-way spectra predicted by the model and their residuals against the data are plotted for the three pairs of populations MSL, GBR (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
+    }
     \label{fig:humans}

### 2025-03-27 example2/plot_fig2.py

@@ -1,0 +1,113 @@
+
+import demes
+import demesdraw
+import matplotlib
+import matplotlib.pylab as plt
+import moments
+import numpy as np
+
+
+plt.rcParams["legend.title_fontsize"] = "xx-small"
+matplotlib.rc("xtick", labelsize=5)
+matplotlib.rc("ytick", labelsize=5)
+matplotlib.rc("axes", labelsize=6)
+matplotlib.rc("axes", titlesize=6)
+matplotlib.rc("legend", fontsize=5)
+
+
+def plot_2d_comp(ax, model, data):
+
+    ax.semilogy(data, "-o", ms=4, lw=1, mfc="w", label="Data")
+    ax.semilogy(model, "-o", ms=2, lw=1, label="Model")        
+    ax.set_xlim(0, data.sample_sizes[0])
+    ax.set_xlabel("Allele frequency")
+    ax.set_ylabel("Count")
+    ax.legend()
+
+    return
+
+
+def main():
+
+    graph = demes.load(
+        "models/MSL_GBR_Vindija_round3/MSL_GBR_Vindija_round3.misid_fit.yaml"
+    )
+    data = moments.Spectrum.from_file("spectra/MSL_GBR_Vindija")
+    p_misid = 0.0223
+    u = 1.5e-8
+    L = 960914001
+    model = moments.Demes.SFS(
+        graph, samples={"MSL": 20, "GBR": 20, "Vindija": 2}, u=u, L=L
+    )
+    model = moments.Misc.flip_ancestral_misid(model, p_misid)
+    resid = moments.Inference.Anscombe_Poisson_residual(model, data)
+
+    grid = (4, 3)
+    fig = plt.figure(1, figsize=(6.5, 5.5))
+    ax0 = plt.subplot2grid(grid, (0, 0), rowspan=2, colspan=2)
+    ax1 = plt.subplot2grid(grid, (0, 2), rowspan=1, colspan=1)
+    ax2 = plt.subplot2grid(grid, (1, 2), rowspan=1, colspan=1)
+    ax3 = plt.subplot2grid(grid, (2, 0), rowspan=1, colspan=1)
+    ax4 = plt.subplot2grid(grid, (2, 1), rowspan=1, colspan=1)
+    ax5 = plt.subplot2grid(grid, (2, 2), rowspan=1, colspan=1)
+    ax6 = plt.subplot2grid(grid, (3, 0), rowspan=1, colspan=1)
[…]

### 2025-03-28 chapter/paper.tex
reorganized example2, improved readmes

@@
 
-We next stitched these one-population models together and refitted the resulting graph to get a simple two-population out-of-Africa model, with continuous symmetric migration between MSL and GBR. 
-Following this, we added a Neandertal branch and fit only the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixture proportion. 
-We also fitted the divergence time between the sampled Vindija Neandertal and the lineage which admixed with modern humans, which we observed to be poorly constrained, likely in part due to the very small size of our Vindija sample.
-In a second round of optimization for the three-population model, we fixed the aforesaid divergence time to a value supported by literature (90 kya, \cite{prufer2017high}) and refit all other parameters to refine the model.
+We next stitched these one-population models together and refitted the resulting graph to get a simple two-population out-of-Africa model that included continuous symmetric migration between MSL and GBR and an ancestral size expansion.
+Following this, we added a Neandertal branch and fitted the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixture proportion. 
+We also fitted the divergence time between the sampled Vindija Neandertal and the Neandertal lineage which admixed with modern humans.
+Finding that this parameter tended to converge to its lower bound (imposed by the estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature (90 kya, \cite{prufer2017high}) in the second round of optimization.
+Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
+Fixing parameters for which good estimates are available in the literature (see e.g.
+\cite{sumer2025earliest}) is often a reasonable thing to do in the first stages of model optimization, and we expected this one to be especially difficult to fit due to the small pulse proportion and Neandertal sample size.
+In a second round of optimization on the three-population model, we refit all the parameters aside from the two times mentioned above to refine the estimates.
 
 After this refitting step, we were left with parameters in good agreement with prior authors, with a Neandertal-modern human divergence time of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 percent.
-We observe some large residuals in our best-fit model, suggesting ...
+We observe some large residuals in our best-fit model (Figure 2), suggesting ...
 

### 2025-03-28 example2/fit_GBR_model.py

@@ -1,0 +1,89 @@
+## Fit a two-epoch model for GBR
+## Then compute confidence intervals and plot the fit
+
+import demes, demesdraw
+import moments
+import matplotlib.pyplot as plt
+
+
+# define model paths
+graph_file = "models/GBR/GBR_model.yaml"
+options_file = "models/GBR/GBR_options.yaml"
+
+# define data path and load data
+data_file = "spectra/GBR"
+data = moments.Spectrum.from_file(data_file)
+
+# define output graph filename
+output_file = "models/GBR/GBR_model.misid_fit.yaml"
+
+# define parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+misid_guess = 0.03
+
+# fit using the Powell algorithm
+param_names, fit_params, ll = moments.Demes.Inference.optimize(
+    graph_file, 
+    options_file, 
+    data, 
+    maxiter=10000,
+    fit_ancestral_misid=True, 
+    misid_guess=misid_guess,
+    uL=U, 
+    verbose=50,
+    output=output_file, 
+    method="powell",
+    overwrite=True
+)
+
+# print results
+print("Best-fit log-likelihood:", -ll)
+print("Best-fit parameters:")
+for name, value in zip(param_names, fit_params):
+    print(f"{name}\t{value:.3}")
+
+# retrieve p_misid
+p_misid = fit_params[-1]
+
+# compute confidence intervals using FIM 
+log_file = "models/GBR/GBR_model.misid_fit.uncerts.txt"
+uncerts = moments.Demes.Inference.uncerts(
+    output_file,
+    options_file,
+    data,
+    uL=U,
+    method="FIM",
+    fit_ancestral_misid=True,
+    misid_fit=p_misid,
+    verbose=10,
+    output=log_file,
+    overwrite=True,
+    eps=1e-3
+)
+
+# print confidence intervals
+print("95% confidence intervals:")
+print("param\t2.5%\t97.5%")
+for name, value, err in zip(param_names, fit_params, uncerts):
+    print(f"{name}\t{value - 1.96 * err:.3}\t{value + 1.96 * err:.3}")
+
+# load output graph
+fit_graph = demes.load(output_file)
+
+# compute model expectations
+model = moments.Demes.SFS(fit_graph, samples={"GBR": 20}, u=u, L=L)
+model = moments.Misc.flip_ancestral_misid(model, p_misid)
+
+# plot size history
[…]

### 2025-03-28 example2/fit_MSL_GBR_Vindija_model_step1.py

@@ -1,0 +1,103 @@
+## Fit parameters to the Neandertal branch of a three-deme MSL, GBR, Vindija model
+## Then compute confidence intervals and plot the fit
+
+import demes, demesdraw
+import moments
+import matplotlib.pyplot as plt
+
+
+# define model paths
+graph_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_model.yaml"
+options_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_options.yaml"
+
+# define data path and load data
+data_file = "spectra/MSL_GBR_Vindija"
+data = moments.Spectrum.from_file(data_file)
+
+# define output graph filename
+output_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_model.misid_fit.yaml"
+
+# define parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+misid_guess = 0.03
+
+# fit using the Powell algorithm
+param_names, fit_params, ll = moments.Demes.Inference.optimize(
+    graph_file, 
+    options_file, 
+    data, 
+    maxiter=10000,
+    fit_ancestral_misid=True, 
+    misid_guess=misid_guess,
+    uL=U, 
+    verbose=1,
+    output=output_file, 
+    method="powell",
+    overwrite=True
+)
+
+# print results
+print("Best-fit log-likelihood:", -ll)
+print("Best-fit parameters:")
+for name, value in zip(param_names, fit_params):
+    print(f"{name}\t{value:.3}")
+
+# retrieve p_misid
+p_misid = fit_params[-1]
+
+# compute confidence intervals using FIM 
+log_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_model.misid_fit.uncerts.txt"
+uncerts = moments.Demes.Inference.uncerts(
+    output_file,
+    options_file,
+    data,
+    uL=U,
+    method="FIM",
+    fit_ancestral_misid=True,
+    misid_fit=p_misid,
+    verbose=10,
+    output=log_file,
+    overwrite=True,
+    eps=1e-2
+)
+
+# print confidence intervals
+print("95% confidence intervals:")
+print("param\t2.5%\t97.5%")
+for name, value, err in zip(param_names, fit_params, uncerts):
[…]

### 2025-03-28 example2/fit_MSL_GBR_Vindija_model_step2.py

@@ -1,0 +1,103 @@
+## Fit parameters to the Neandertal branch of a three-deme MSL, GBR, Vindija model
+## Then compute confidence intervals and plot the fit
+
+import demes, demesdraw
+import moments
+import matplotlib.pyplot as plt
+
+
+# define model paths
+graph_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_model_step2.yaml"
+options_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_options_step2.yaml"
+
+# define data path and load data
+data_file = "spectra/MSL_GBR_Vindija"
+data = moments.Spectrum.from_file(data_file)
+
+# define output graph filename
+output_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_model_step2.misid_fit.yaml"
+
+# define parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+misid_guess = 0.03
+
+# fit using the Powell algorithm
+param_names, fit_params, ll = moments.Demes.Inference.optimize(
+    graph_file, 
+    options_file, 
+    data, 
+    maxiter=10000,
+    fit_ancestral_misid=True, 
+    misid_guess=misid_guess,
+    uL=U, 
+    verbose=1,
+    output=output_file, 
+    method="powell",
+    overwrite=True
+)
+
+# print results
+print("Best-fit log-likelihood:", -ll)
+print("Best-fit parameters:")
+for name, value in zip(param_names, fit_params):
+    print(f"{name}\t{value:.3}")
+
+# retrieve p_misid
+p_misid = fit_params[-1]
+
+# compute confidence intervals using FIM 
+log_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_model_step2.misid_fit.uncerts.txt"
+uncerts = moments.Demes.Inference.uncerts(
+    output_file,
+    options_file,
+    data,
+    uL=U,
+    method="FIM",
+    fit_ancestral_misid=True,
+    misid_fit=p_misid,
+    verbose=10,
+    output=log_file,
+    overwrite=True,
+    eps=1e-2
+)
+
+# print confidence intervals
+print("95% confidence intervals:")
+print("param\t2.5%\t97.5%")
[…]

### 2025-03-28 example2/fit_MSL_GBR_model.py

@@ -1,0 +1,101 @@
+## Fit a three-epoch model for MSL_GBR
+## Then compute confidence intervals and plot the fit
+
+import demes, demesdraw
+import moments
+import matplotlib.pyplot as plt
+
+
+# define model paths
+graph_file = "models/MSL_GBR/MSL_GBR_model.yaml"
+options_file = "models/MSL_GBR/MSL_GBR_options.yaml"
+
+# define data path and load data
+data_file = "spectra/MSL_GBR"
+data = moments.Spectrum.from_file(data_file)
+
+# define output graph filename
+output_file = "models/MSL_GBR/MSL_GBR_model.misid_fit.yaml"
+
+# define parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+misid_guess = 0.03
+
+# fit using the fmin algorithm
+param_names, fit_params, ll = moments.Demes.Inference.optimize(
+    graph_file, 
+    options_file, 
+    data, 
+    maxiter=10000,
+    fit_ancestral_misid=True, 
+    misid_guess=misid_guess,
+    uL=U, 
+    verbose=1,
+    output=output_file, 
+    method="fmin",
+    overwrite=True
+)
+
+# print results
+print("Best-fit log-likelihood:", -ll)
+print("Best-fit parameters:")
+for name, value in zip(param_names, fit_params):
+    print(f"{name}\t{value:.3}")
+
+# retrieve p_misid
+p_misid = fit_params[-1]
+
+# compute confidence intervals using FIM 
+log_file = "models/MSL_GBR/MSL_GBR_model.misid_fit.uncerts.txt"
+uncerts = moments.Demes.Inference.uncerts(
+    output_file,
+    options_file,
+    data,
+    uL=U,
+    method="FIM",
+    fit_ancestral_misid=True,
+    misid_fit=p_misid,
+    verbose=10,
+    output=log_file,
+    overwrite=True,
+    eps=1e-3
+)
+
+# print confidence intervals
+print("95% confidence intervals:")
+print("param\t2.5%\t97.5%")
+for name, value, err in zip(param_names, fit_params, uncerts):
+    print(f"{name}\t{value - 1.96 * err:.3}\t{value + 1.96 * err:.3}")
+
+# load output graph
+fit_graph = demes.load(output_file)
+
+# compute model expectations
[…]

### 2025-03-28 example2/fit_MSL_model.py

@@ -1,0 +1,95 @@
+## Fit a one-population, three-epoch model for MSL
+## Then compute confidence intervals and plot the fit
+
+import demes, demesdraw
+import moments
+import matplotlib.pyplot as plt
+
+
+# define model paths
+graph_file = "models/MSL/MSL_model.yaml"
+options_file = "models/MSL/MSL_options.yaml"
+
+# define data path and load data
+data_file = "spectra/MSL"
+data = moments.Spectrum.from_file(data_file)
+
+# define output graph filename
+output_file = "models/MSL/MSL_model.misid_fit.yaml"
+
+# define parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+misid_guess = 0.03
+
+# fit using the Powell algorithm
+param_names, fit_params, ll = moments.Demes.Inference.optimize(
+    graph_file, 
+    options_file, 
+    data, 
+    uL=U, 
+    fit_ancestral_misid=True, 
+    misid_guess=misid_guess,
+    verbose=50,
+    output=output_file, 
+    maxiter=10000,
+    method="powell",
+    overwrite=True
+)
+
+# print results
+print("Best-fit log-likelihood:", -ll)
+print("Best-fit parameters:")
+for name, value in zip(param_names, fit_params):
+    print(f"{name}\t{value:.3}")
+
+# retrieve p_misid
+p_misid = fit_params[-1]
+
+# compute confidence intervals using FIM 
+log_file = "models/MSL/MSL_model.misid_fit.uncerts.txt"
+uncerts = moments.Demes.Inference.uncerts(
+    output_file,
+    options_file,
+    data,
+    uL=U,
+    fit_ancestral_misid=True,
+    misid_fit=p_misid,
+    verbose=10,
+    method="FIM",
+    output=log_file,
+    overwrite=True,
+    eps=1e-3
+)
+
+# print uncerts
+print("Uncerts:")
+print("param\tstd_err")
+for name, err in zip(param_names, uncerts):
+    print(f"{name}\t{err:.3}")
+
+# print confidence intervals
+print("95% confidence intervals:")
+print("param\t2.5%\t97.5%")
+for name, value, err in zip(param_names, fit_params, uncerts):
+    print(f"{name}\t{value - 1.96 * err:.3}\t{value + 1.96 * err:.3}")
+
+# load output graph
+fit_graph = demes.load(output_file)
+
[…]

### 2025-03-28 example2/models/GBR/fit_model.py

@@ -1,40 +1,0 @@
-## Fit a one-population model for GBR
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "GBR_model.yaml"
-    options_file = "options_GBR.yaml"
-    data_file = "../../spectra/GBR"
-    output = "GBR_model.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-
-    data = moments.Spectrum.from_file(data_file)
-
-    # fit using the Powell algorithm
-    param_names, fit_params, ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=0.03,
-        uL=u * L, 
-        verbose=50,
-        output=output, 
-        method="powell"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -ll)
-    print("Best-fit parameters:")
-    for name, val in zip(param_names, fit_params):
-        print(f"{name}\t{val:.3}")
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/GBR/plot_model.py

@@ -1,35 +1,0 @@
-## Make a plot of expectations from the model fit by ./fit_model.py
-
-import demes
-import demesdraw
-import matplotlib.pyplot as plt
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "GBR_model.misid_fit.yaml"
-    data_file = "../../spectra/GBR"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0576934
-
-    graph = demes.load(graph_file)
-    model = moments.Demes.SFS(graph, samples={"GBR": 20}, u=u, L=L)
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-    data = moments.Spectrum.from_file(data_file)
-
-    # plot size history
-    demesdraw.size_history(graph)
-    plt.savefig("GBR_model.misid_fit.size_history.png")
-    plt.close()
-
-    # plot SFS fit
-    moments.Plotting.plot_1d_comp_Poisson(
-        model, data, residual="linear", out="GBR_model.misid_fit.comp_1d.png"
-    )
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL/fit_model.py

@@ -1,41 +1,0 @@
-## Fit a one-population model for MSL
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_model.yaml"
-    options_file = "options_MSL.yaml"
-    data_file = "../../spectra/MSL"
-    output = "MSL_model.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-    misid_guess = 0.03
-
-    data = moments.Spectrum.from_file(data_file)
-
-    # fit using the Powell algorithm
-    param_names, fit_params, ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=misid_guess,
-        uL=u * L, 
-        verbose=50,
-        output=output, 
-        method="powell"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -ll)
-    print("Best-fit parameters:")
-    for name, val in zip(param_names, fit_params):
-        print(f"{name}\t{val:.3}")
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL/fit_model_folded.py

@@ -1,42 +1,0 @@
-## Fit a one-population model for MSL using the folded SFS
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_model.yaml"
-    options_file = "options_MSL.yaml"
-    data_file = "../../spectra/MSL"
-    output = "MSL_model.folded_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-
-    data = moments.Spectrum.from_file(data_file)
-    
-    # fold the SFS
-    data = data.fold()
-
-    # fit using the Powell algorithm
-    param_names, fit_params, ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=0.03,
-        uL=u * L, 
-        verbose=50,
-        output=output, 
-        method="powell"
-    )
-    # print results
-    print("Best-fit log-likelihood:", -ll)
-    print("Best-fit parameters:")
-    for name, val in zip(param_names, fit_params):
-        print(f"{name}\t{val:.3}")
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL/plot_model.py

@@ -1,37 +1,0 @@
-## Make a plot of expectations from the model fit by ./fit_model.py
-
-import demes
-import demesdraw
-import matplotlib.pyplot as plt
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_model.misid_fit.yaml"
-    data_file = "../../spectra/MSL"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0318905
-
-    # load data and obtain model expectations (adjusted by fit misid pr.)
-    graph = demes.load(graph_file)
-    model = moments.Demes.SFS(graph, samples={"MSL": 20}, u=u, L=L)
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-    data = moments.Spectrum.from_file(data_file)
-
-    # plot size history
-    ax = plt.subplot(111)
-    demesdraw.size_history(graph, ax=ax)
-    plt.savefig("MSL_model.misid_fit.size_history.png")
-    plt.close()
-
-    # plot SFS fit
-    moments.Plotting.plot_1d_comp_Poisson(
-        model, data, residual="linear", out="MSL_model.misid_fit.comp_1d.png"
-    )
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL/plot_model_folded.py

@@ -1,36 +1,0 @@
-## Make a plot of expectations from the model fit by ./fit_model_folded.py
-
-import demes
-import demesdraw
-import matplotlib.pyplot as plt
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_model.folded_fit.yaml"
-    data_file = "../../spectra/MSL"
-    u = 1.5e-8
-    L = 960914001
-
-    # load data, obtain model expectations and fold them
-    graph = demes.load(graph_file)
-    model = moments.Demes.SFS(graph, samples={"MSL": 20}, u=u, L=L)
-    model = model.fold()
-    data = moments.Spectrum.from_file(data_file)
-    data = data.fold()
-
-    # plot size history
-    demesdraw.size_history(graph)
-    plt.savefig("MSL_model.folded_fit.size_history.png")
-    plt.close()
-
-    # plot SFS fit
-    moments.Plotting.plot_1d_comp_Poisson(
-        model, data, residual="linear", out="MSL_model.folded_fit.comp_1d.png"
-    )
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL_GBR/fit_model.py

@@ -1,42 +1,0 @@
-## Fit a model of MSL and GBR populations.
-
-import moments 
-import sys
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_model.yaml"
-    options_file = "options_MSL_GBR.yaml"
-    data_file = "../../spectra/MSL_GBR"
-    output = "MSL_GBR_model.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-    misid_guess = 0.035
-
-    data = moments.Spectrum.from_file(data_file)
-
-    # fit model using the Powell algorithm
-    param_names, fit_params, ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=misid_guess,
-        uL=u * L, 
-        verbose=10,
-        output=output, 
-        method="powell"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -ll)
-    print("Best-fit parameters:")
-    for name, val in zip(param_names, fit_params):
-        print(f"{name}\t{val:.3}")
-
-    return 
-
-
-main() 

### 2025-03-28 example2/models/MSL_GBR/plot_model.py

@@ -1,56 +1,0 @@
-## Make a plot of expectations from the model fit by ./fit_model.py
-
-import demes
-import demesdraw
-import matplotlib.pyplot as plt
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_model.misid_fit.yaml"
-    data_file = "../../spectra/MSL_GBR"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0308139
-
-    # load data and obtain model expectations (adjusted by fit misid pr.)
-    graph = demes.load(graph_file)
-    model = moments.Demes.SFS(graph, samples={"MSL": 20, "GBR": 20}, u=u, L=L)
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-    data = moments.Spectrum.from_file(data_file)
-
-    # plot size history
-    ax = plt.subplot(111)
-    demesdraw.size_history(graph, ax=ax)
-    plt.savefig("MSL_GBR_model.misid_fit.size_history.png")
-    plt.close()
-
-    # plot marginal fit for MSL
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([1]), 
-        data.marginalize([1]), 
-        residual="linear", 
-        out="MSL_GBR_model.misid_fit.comp_1d_MSL.png"
-    )
-
-    # plot marginal fit for GBR
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([0]), 
-        data.marginalize([0]), 
-        residual="linear", 
-        out="MSL_GBR_model.misid_fit.comp_1d_GBR.png"
-    )
-
-    # plot joint fit (2d) for MSL, GBR
-    moments.Plotting.plot_2d_comp_Poisson(
-        model, 
-        data, 
-        residual="linear", 
-        out="MSL_GBR_model.misid_fit.comp_2d.png"
-    )
-
-    return 
-
-    
-main() 

### 2025-03-28 example2/models/MSL_GBR_Vindija_round1/fit_model.py

@@ -1,41 +1,0 @@
-## Fit a model of MSL, GBR, and Vindija populations.
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_Vindija_model.yaml"
-    options_file = "options_MSL_GBR_Vindija.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    output = "MSL_GBR_Vindija_model.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-    misid_guess = 0.03
-
-    data = moments.Spectrum.from_file(data_file)
-
-    # fit using the LBFGSB algorithm
-    param_names, fit_params, ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=misid_guess,
-        uL=u * L, 
-        verbose=1,
-        output=output, 
-        method="lbfgsb"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -ll)
-    print("Best-fit parameters:")
-    for name, val in zip(param_names, fit_params):
-        print(f"{name}\t{val:.3}")
-
-    return 
-
-
-main() 

### 2025-03-28 example2/models/MSL_GBR_Vindija_round1/plot_model.py

@@ -1,57 +1,0 @@
-## Make model/data comparison plots for the model fit by ./fit_model.py
-
-import demes
-import demesdraw 
-import matplotlib.pyplot as plt
-import moments
-
-
-def main():
-
-    graph_file = "MSL_GBR_Vindija_model.misid_fit.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0250059
-
-    graph = demes.load(graph_file)
-    data = moments.Spectrum.from_file(data_file)
-    model = moments.Demes.SFS(
-        graph, 
-        samples={"MSL":20, "GBR":20, "Vindija": 2}, 
-        u=u,
-        L=L
-    )
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-
-    # plot demography in linear and log time
-    demesdraw.tubes(graph, log_time=True)
-    plt.savefig("MSL_GBR_Vindija_model.misid_fit.tubes_logtime.png")
-    plt.close()
-
-    demesdraw.tubes(graph)
-    plt.savefig("MSL_GBR_Vindija_model.misid_fit.tubes.png")
-    plt.close()
-
-    # plot marginal fit for GBR
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([0, 2]), 
-        data.marginalize([0, 2]),
-        residual="linear", 
-        out="MSL_GBR_Vindija_model.misid_fit.comp_1d_GBR.png"
-    )   
-
-    # plot 3d comparison
-    moments.Plotting.plot_3d_comp_Poisson(
-        model, 
-        data, 
-        residual="linear", 
-        out="MSL_GBR_Vindija_model.misid_fit.comp_3d.png"
-    )
-
-    return
-
-
-if __name__ == "__main__":
-    main()
-

### 2025-03-28 example2/models/MSL_GBR_Vindija_round2/fit_model.py

@@ -1,40 +1,0 @@
-## Refit the MSL, GBR, Vindija model to fine-tune parameters
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_Vindija_round2.yaml"
-    options_file = "options_MSL_GBR_Vindija_round2.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    output = "MSL_GBR_Vindija_round2.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-    misid_guess = 0.024
-
-    data = moments.Spectrum.from_file(data_file)
-
-    param_names, misid_popt, misid_ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=misid_guess,
-        uL=u * L, 
-        verbose=1,
-        output=output,
-        method="lbfgsb"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -misid_ll)
-    print("Best-fit parameters:")
-    for name, popt in zip(param_names, misid_popt):
-        print(f"{name}\t{popt:.3}")
-
-    return 
-
-
-main() 

### 2025-03-28 example2/models/MSL_GBR_Vindija_round2/plot_model.py

@@ -1,67 +1,0 @@
-## Make model/data comparison plots for the model fit by ./fit_model.py
-
-import demes
-import demesdraw 
-import matplotlib.pyplot as plt
-import moments
-
-
-def main():
-
-    graph_file = "MSL_GBR_Vindija_round2.misid_fit.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0218
-
-    graph = demes.load(graph_file)
-    data = moments.Spectrum.from_file(data_file)
-    model = moments.Demes.SFS(
-        graph, 
-        samples={"MSL": 20, "GBR": 20, "Vindija": 2}, 
-        u=u,
-        L=L
-    )
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-
-    print(moments.Inference.ll(model, data))
-
-    # plot demography in linear and log time
-    ax = plt.subplot(111)
-    demesdraw.tubes(graph, log_time=True, ax=ax)
-    plt.savefig("MSL_GBR_Vindija_round2.misid_fit.tubes_logtime.png")
-    plt.close()
-    ax = plt.subplot(111)
-    demesdraw.tubes(graph, ax=ax)
-    plt.savefig("MSL_GBR_Vindija_round2.misid_fit.tubes.png")
-    plt.close()
-
-    # plot marginal fit for GBR
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([0, 2]), 
-        data.marginalize([0, 2]),
-        residual="linear", 
-        out="MSL_GBR_Vindija_round2.misid_fit.comp_1d_GBR.png"
-    )
-    # and for MSL
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([1, 2]), 
-        data.marginalize([1, 2]),
-        residual="linear", 
-        out="MSL_GBR_Vindija_round2.misid_fit.comp_1d_MSL.png"
-    )
-
-    # plot 3d comparison
-    moments.Plotting.plot_3d_comp_Poisson(
-        model, 
-        data, 
-        residual="linear", 
-        out="MSL_GBR_Vindija_round2.misid_fit.comp_3d.png"
-    )
-
-    return
-
-
-if __name__ == "__main__":
-    main()
-

### 2025-03-28 example2/models/MSL_GBR_Vindija_round3/compute_uncerts.py

@@ -1,42 +1,0 @@
-## Compute uncertainties using the FIM method
-
-import demes
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_Vindija_round3.misid_fit.yaml"
-    options_file = "options_MSL_GBR_Vindija_round3.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0223
-
-    data = moments.Spectrum.from_file(data_file)
-
-    # load parameter names and best-fit values using moments.Demes functions
-    builder = demes.load(graph_file).asdict()
-    options = moments.Demes.Inference._get_params_dict(options_file)
-    param_names, fit_params, _, __ = \
-        moments.Demes.Inference._set_up_params_and_bounds(options, builder)
-
-    std_errs = moments.Demes.Inference.uncerts(
-        graph_file,
-        options_file,
-        data,
-        uL=u * L,
-        fit_ancestral_misid=True,
-        misid_fit=p_misid,
-    )
-
-    # print results
-    print(r"95% confidence intervals:")
-    print("param\t2.5%\t97.5%")
-    for name, val, err in zip(param_names, fit_params, std_errs):
-        print(f"{name}\t{val - 1.96 * err:.3}\t{val + 1.96 * err:.3}")
-
-    return 
-
-
-main() 

### 2025-03-28 example2/models/MSL_GBR_Vindija_round3/fit_model.py

@@ -1,40 +1,0 @@
-## Refit the MSL, GBR, Vindija model to fine-tune parameters
-
-import moments 
-
-
-def main():
-    # define paths and parameters
-    graph_file = "MSL_GBR_Vindija_round3.yaml"
-    options_file = "options_MSL_GBR_Vindija_round3.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    output = "MSL_GBR_Vindija_round3.misid_fit.yaml"
-    u = 1.5e-8
-    L = 960914001
-    misid_guess = 0.024
-
-    data = moments.Spectrum.from_file(data_file)
-
-    param_names, misid_popt, misid_ll = moments.Demes.Inference.optimize(
-        graph_file, 
-        options_file, 
-        data, 
-        maxiter=10000,
-        fit_ancestral_misid=True, 
-        misid_guess=misid_guess,
-        uL=u * L, 
-        verbose=1,
-        output=output,
-        method="lbfgsb"
-    )
-
-    # print results
-    print("Best-fit log-likelihood:", -misid_ll)
-    print("Best-fit parameters:")
-    for name, popt in zip(param_names, misid_popt):
-        print(f"{name}\t{popt:.3}")
-
-    return 
-
-
-main() 

### 2025-03-28 example2/models/MSL_GBR_Vindija_round3/plot_model.py

@@ -1,67 +1,0 @@
-## Make model/data comparison plots for the model fit by ./fit_model.py
-
-import demes
-import demesdraw 
-import matplotlib.pyplot as plt
-import moments
-
-
-def main():
-
-    graph_file = "MSL_GBR_Vindija_round3.misid_fit.yaml"
-    data_file = "../../spectra/MSL_GBR_Vindija"
-    u = 1.5e-8
-    L = 960914001
-    p_misid = 0.0223
-
-    graph = demes.load(graph_file)
-    data = moments.Spectrum.from_file(data_file)
-    model = moments.Demes.SFS(
-        graph, 
-        samples={"MSL": 20, "GBR": 20, "Vindija": 2}, 
-        u=u,
-        L=L
-    )
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-
-    print(moments.Inference.ll(model, data))
-
-    # plot demography in linear and log time
-    ax = plt.subplot(111)
-    demesdraw.tubes(graph, log_time=True, ax=ax)
-    plt.savefig("MSL_GBR_Vindija_round3.misid_fit.tubes_logtime.png")
-    plt.close()
-    ax = plt.subplot(111)
-    demesdraw.tubes(graph, ax=ax)
-    plt.savefig("MSL_GBR_Vindija_round3.misid_fit.tubes.png")
-    plt.close()
-
-    # plot marginal fit for GBR
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([0, 2]), 
-        data.marginalize([0, 2]),
-        residual="linear", 
-        out="MSL_GBR_Vindija_round3.misid_fit.comp_1d_GBR.png"
-    )
-    # and for MSL
-    moments.Plotting.plot_1d_comp_Poisson(
-        model.marginalize([1, 2]), 
-        data.marginalize([1, 2]),
-        residual="linear", 
-        out="MSL_GBR_Vindija_round3.misid_fit.comp_1d_MSL.png"
-    )
-
-    # plot 3d comparison
-    moments.Plotting.plot_3d_comp_Poisson(
-        model, 
-        data, 
-        residual="linear", 
-        out="MSL_GBR_Vindija_round3.misid_fit.comp_3d.png"
-    )
-
-    return
-
-
-if __name__ == "__main__":
-    main()
-

### 2025-03-28 example2/plot_fig2.py

@@ -27,87 +27,81 @@ def plot_2d_comp(ax, model, data):
     return
 
 
-def main():
+graph = demes.load(
+    "models/MSL_GBR_Vindija_round3/MSL_GBR_Vindija_round3.misid_fit.yaml"
+)
+data = moments.Spectrum.from_file("spectra/MSL_GBR_Vindija")
+p_misid = 0.0223
+u = 1.5e-8
+L = 960914001
+model = moments.Demes.SFS(
+    graph, samples={"MSL": 20, "GBR": 20, "Vindija": 2}, u=u, L=L
+)
+model = moments.Misc.flip_ancestral_misid(model, p_misid)
+resid = moments.Inference.Anscombe_Poisson_residual(model, data)
 
-    graph = demes.load(
-        "models/MSL_GBR_Vindija_round3/MSL_GBR_Vindija_round3.misid_fit.yaml"
-    )
-    data = moments.Spectrum.from_file("spectra/MSL_GBR_Vindija")
-    p_misid = 0.0223
-    u = 1.5e-8
-    L = 960914001
-    model = moments.Demes.SFS(
-        graph, samples={"MSL": 20, "GBR": 20, "Vindija": 2}, u=u, L=L
-    )
-    model = moments.Misc.flip_ancestral_misid(model, p_misid)
-    resid = moments.Inference.Anscombe_Poisson_residual(model, data)
+grid = (4, 3)
+fig = plt.figure(1, figsize=(6.5, 5.5))
+ax0 = plt.subplot2grid(grid, (0, 0), rowspan=2, colspan=2)
+ax1 = plt.subplot2grid(grid, (0, 2), rowspan=1, colspan=1)
+ax2 = plt.subplot2grid(grid, (1, 2), rowspan=1, colspan=1)
+ax3 = plt.subplot2grid(grid, (2, 0), rowspan=1, colspan=1)
+ax4 = plt.subplot2grid(grid, (2, 1), rowspan=1, colspan=1)
+ax5 = plt.subplot2grid(grid, (2, 2), rowspan=1, colspan=1)
+ax6 = plt.subplot2grid(grid, (3, 0), rowspan=1, colspan=1)
+ax7 = plt.subplot2grid(grid, (3, 1), rowspan=1, colspan=1)
+ax8 = plt.subplot2grid(grid, (3, 2), rowspan=1, colspan=1)
 
-    grid = (4, 3)
[…]

### 2025-03-31 chapter/paper.tex
small revisions to chapter

@@
 \begin{abstract}
-    Placeholder
+In this chapter we show how to use \moments, a Python library for computing the
+joint SFS and inferring demographic history, through its \demes interface. 
+Following a discussion of SFS inference, we illustrate the general usage of 
+\moments and sketch two specific examples. We conclude with a discussion of some 
+additional considerations regarding the \moments library.
 \end{abstract}
@@
 the genome and how those gene-genealogical relationships change along a
-chromosome due to recombination (that is, information stored in the Ancestral
+chromosome due to recombination (that is, by information stored in the Ancestral
 Recombination Graph \citep{nielsen2025inference}). However, the ARG can be
@@
 \citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
-code across software, and ease of model implementation and reduction of errors.
+code across software, ease of model implementation and reduction of errors.
 
@@
 joint SFS. The first example simulates data with known demographic parameters,
-which we try to reinfer using the both the original and misspecified simpler
-models. In the second example, we infer a relatively simple human-Neanderthal
+which we try to reinfer using the both the original model and simpler misspecified 
+ones. In the second example, we infer a relatively simple human-Neanderthal
 history using data from two human populations and a high-coverage ancient
 sample from the Neanderthal lineage. For both of these examples, we briefly
-describe data processing and inference setup, highlighting the high-level
-components of the analyses. For detailed information for each example,
+describe data processing and inference setup, emphasizing the high-level
+components of the analyses. For detailed information about each example,
 including managing data, specifying models, performing inference, computing
@@
 which parameters should be fit. For this, we use a separate YAML-formatted file
-to define \texttt{parameters}, each of which points to the value(s) in the
+to define \texttt{parameters}, each of which points to desired value(s) in the
 demographic model and sets lower and upper bounds. This file also specifies any
@@
 we have an estimate for the total mutation rate, $U$.
-\begin{minted}[gobble=4, frame=single]{YAML}
+\begin{minted}[gobble=4, frame=single]{python}
     import moments
@@
 
-Confidence intervals may also be computed, once a (locally) optimal model has
-been found.
+Confidence intervals may also be computed using 
+\texttt{moments.Demes.Inference.uncerts(...)}, once a (locally) optimal model 
+has been found. The following will emit a tab-separated file containing
+parameter names, best-fit values and estimated standard errors.
+\begin{minted}[gobble=4, frame=single]{python}
+    log_file = 'log.txt'
+
+    uncerts = moments.Demes.Inference.uncerts(
+      output, params_file, data,
+      uL=U, output=log_file
+    )
+\end{minted}
 
@@
 
-We obtained modern human genome sequences from a recent resequencing of the 1000 Genomes Project cohort \citep{byrska2022high}, and a high-coverage sequence of the Vindija Neandertal from \cite{prufer2017high}.
-From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR (British from England and Scotland) sequences as our modern human samples.  
-After filtering out sites that lay outside the 1000 Genomes `strict' mask, that were not genotyped in all four existing high-coverage archaic genome sequences, lacked high-confidence ancestral state assignments, or fell within 10 kb of exonic or promoter regions, we were left with approximately 960 Mb of well-characterized and putatively neutrally-evolving sequence.
+We obtained modern human genome sequences from a recent resequencing of the 
+1000 Genomes Project cohort \citep{byrska2022high}, and used the high-coverage 
+sequence of the Vindija Neandertal from \cite{prufer2017high}. 
+From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
+GBR (British from England and Scotland) sequences as our modern human samples.  
+We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
+were not genotyped in all four existing high-coverage archaic genome sequences, 
+thatlacked high-confidence ancestral state assignments, or fell within 10 kb of 
+exonic or promoter regions and were left with approximately 960 Mb of 
+well-characterized and putatively neutrally-evolving sequence.
 We estimated the genome-wide SFS using built-in \moments parsing functions.
-Details concerning data processing (including our liftover of the Vindija sequence to a more recent genome build) can be found in the linked GitHub repository.
+Details concerning data processing (including our liftover of the Vindija 
+sequence to a more recent genome build) can be found in the linked GitHub 
+repository.
 
-The model that we wished to fit is relatively simple but still has many parameters (11 in the example we present), so we constructed it in stages by first fitting simpler one- and two-population graphs.
-Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per generation and 29 years as the average generation time.
-We began by fitting one-population models with ancestral and recent size changes for MSL and GBR.
-For the MSL model, we allowed three epochs with constant effective size in each epoch.
-We attributed to GBR a sharp contraction $\sim$60 kya followed by exponential growth to represent the out-of-Africa bottleneck and subsequent rapid expansion.
-YAML files representing the models and parameters in this section can be found in the linked GitHub repository.
+The model that we wished to fit is relatively simple but still has many 
+parameters (11 in the example we present), so we constructed it in stages by 
+first fitting simpler one-- and two--population models.
+Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per 
+generation and 29 years as the average generation time. We began by fitting 
+one-population models with ancestral and recent size changes for MSL and GBR.
+For the MSL model, we allowed three epochs with constant effective size in each 
+epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
+exponential growth to represent the out-of-Africa bottleneck and subsequent 
+rapid expansion. YAML files representing the models and parameters in this 
+section can be found in the linked GitHub repository.
 
-We next stitched these one-population models together and refitted the resulting graph to get a simple two-population out-of-Africa model that included continuous symmetric migration between MSL and GBR and an ancestral size expansion.
-Following this, we added a Neandertal branch and fitted the Neandertal-modern human divergence time, Neandertal effective size, and the Neandertal-to-GBR admixture proportion. 
-We also fitted the divergence time between the sampled Vindija Neandertal and the Neandertal lineage which admixed with modern humans.
-Finding that this parameter tended to converge to its lower bound (imposed by the estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature (90 kya, \cite{prufer2017high}) in the second round of optimization.
+We next stitched these one-population models together and refitted the 
+resulting graph to get a two-population model that included continuous symmetric 
+migration between MSL and GBR and an ancestral size expansion. Following this,
+we added a Neandertal branch and fitted the 
+Neandertal-modern human divergence time, Neandertal effective size, and the 
+Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
+between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
+admixed with modern non-African humans. Finding that this parameter tended to 
+converge to its lower bound (imposed by the estimated time of death of the 
+Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
+(90 kya, \cite{prufer2017high}) in a second round of optimization.
 Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
-Fixing parameters for which good estimates are available in the literature (see e.g.
-\cite{sumer2025earliest}) is often a reasonable thing to do in the first stages of model optimization, and we expected this one to be especially difficult to fit due to the small pulse proportion and Neandertal sample size.
-In a second round of optimization on the three-population model, we refit all the parameters aside from the two times mentioned above to refine the estimates.
+Fixing parameters for which good estimates are available in the literature 
+(see e.g. \cite{sumer2025earliest}) may be reasonable in the first stages of 
+model optimization, and we expected this value to be especially difficult to fit 
+due to the small pulse proportion and Neandertal sample size. In the second 
+round of optimization on the three-population model, we refit all parameters 
+aside from the two event times mentioned above to refine the estimates.
 
-After this refitting step, we were left with parameters in good agreement with prior authors, with a Neandertal-modern human divergence time of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 percent.
-We observe some large residuals in our best-fit model (Figure 2), suggesting ...
+After this refitting step, we were left with parameters in good agreement with 
+prior authors, with a Neandertal-modern human divergence time of 547 kya, 
+MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
+percent. We observe some large residuals in our best-fit model 
+Figure~\ref{fig:humans}G-I), suggesting ...
 
@@
       A plot of the best-fit model created using \demesdraw is displayed in (A).
-      Marginal SFS fits between the model and empirical data are shown for MSL (B) and GBR (C).
-      Marginal two-way spectra predicted by the model and their residuals against the data are plotted for the three pairs of populations MSL, GBR (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
+      Marginal SFS fits between the model and empirical data are shown for MSL 
+      (B) and GBR (C).
+      Marginal two-way spectra predicted by the model and their residuals 
+      against the data are plotted for the three pairs of populations MSL, GBR 
+      (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
     }

### 2025-04-01 chapter/paper.tex
add finished draft

@@
 \usepackage{color}
+\usepackage{setspace}
 
@@
 
+%\onehalfspacing
+
 \begin{abstract}
-In this chapter we show how to use \moments, a Python library for computing the
-joint SFS and inferring demographic history, through its \demes interface. 
-Following a discussion of SFS inference, we illustrate the general usage of 
-\moments and sketch two specific examples. We conclude with a discussion of some 
-additional considerations regarding the \moments library.
+
+    The site frequency spectrum (SFS) describes the distribution of allele
+    frequencies in one or more populations and sees wide use in evolutionary
+    inference. Here we show how to use \moments, a Python library for working
+    with the SFS and other measures of genetic diversity, and \demes, a
+    standard format for parameterizing population models, to infer demographic
+    history. Following an introduction to SFS inference, we illustrate the
+    general usage of \moments and its \demes interface, and then show two
+    specific examples, using both simulated and human genetic data. We conclude
+    with caveats and considerations regarding the \moments library and SFS
+    inference more broadly.
+
 \end{abstract}
@@
 the genome and how those gene-genealogical relationships change along a
-chromosome due to recombination (that is, by information stored in the Ancestral
-Recombination Graph \citep{nielsen2025inference}). However, the ARG can be
-large and unwieldy, and methods for reconstructing history directly from the
-ARG, while showing promise \citep[e.g.,][]{yc2022evaluation, fan2023likelihood,
-brandt2024promise}, are in their infancy and so far limited in application and
-scalability. Instead, evolutionary inference using informative summaries of
-genetic variation remains a tractable and powerful alternative for learning
-parameters of population history, natural selection and genome biology.
+chromosome due to recombination (that is, by information stored in the
+Ancestral Recombination Graph \citep{nielsen2025inference}). However, the ARG
+can be large and unwieldy, and methods for reconstructing history directly from
+the ARG, while showing promise \citep[e.g.,][]{yc2022evaluation,
+fan2023likelihood, brandt2024promise}, are in their infancy and so far limited
+in application and scalability. Instead, evolutionary inference using
+informative summaries of genetic variation remains a tractable and powerful
+alternative for learning parameters of population history, natural selection
+and genome biology.
 
@@
 and inference (including \msprime \citep{baumdicker2022efficient}, \momi
-\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, \texttt{GADMA}
-\citep{noskova2023gadma2}). \demes encourages interoperability and reuse of
-code across software, ease of model implementation and reduction of errors.
+\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, and
+\texttt{GADMA} \citep{noskova2023gadma2}). \demes encourages interoperability
+and reuse of code across software, ease of model implementation and reduction
+of errors.
 
 Demographic models are specified in YAML format (see \citet{gower2022demes} and
-the associated documentation). The two-population split-with-migration model
-depicted in Figure~\ref{fig:im}A is specified as below:
+the associated documentation). To illustrate, the two-population
+split-with-migration model depicted in Figure~\ref{fig:im}A is specified as
+below:
 \begin{minted}[gobble=4, frame=single,]{yaml}
@@
 \citep{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
-to load the demographic model using \demes and specify the number of haploid
-samples to draw and from which populations.
+to load the demographic model using \demes and then specify the number of
+haploid samples to draw and from which populations.
 \begin{minted}[gobble=4, frame=single]{python}
@@
 joint SFS. The first example simulates data with known demographic parameters,
-which we try to reinfer using the both the original model and simpler misspecified 
-ones. In the second example, we infer a relatively simple human-Neanderthal
-history using data from two human populations and a high-coverage ancient
-sample from the Neanderthal lineage. For both of these examples, we briefly
-describe data processing and inference setup, emphasizing the high-level
-components of the analyses. For detailed information about each example,
-including managing data, specifying models, performing inference, computing
-confidence intervals and visualizing fits, we refer readers to the accompanying
-GitHub repository
+which we try to reinfer using the both the original model and simpler
+misspecified models. In the second example, we infer a relatively simple
+human-Neanderthal history using data from two human populations and a
+high-coverage ancient sample from the Neanderthal lineage. For both of these
+examples, we briefly describe data processing and inference setup, emphasizing
+the high-level components of the analyses. For detailed information about each
+example, including managing data, specifying models, performing inference,
+computing confidence intervals and visualizing fits, we refer readers to the
+accompanying GitHub repository
 (\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments}), which
 is maintained with up-to-date versioning.
-
-\begin{figure}[t!]
-    \includegraphics{../example1/fig1.pdf}
-    \caption{Caption placeholder}
-    \label{fig:im}
-\end{figure}
 
@@
 to define \texttt{parameters}, each of which points to desired value(s) in the
-demographic model and sets lower and upper bounds. This file also specifies any
-relative \texttt{constraints} between pairs of parameters, to ensure the
-optimization routine only explores valid model space. Below, we have included
-a truncated parameters file -- the full example is available on GitHub.
+demographic model and for which lower and upper bounds may be set. This file
+also specifies any relative \texttt{constraints} between pairs of parameters,
+to ensure the optimization routine only explores valid model space. Below, we
+have included a truncated parameters file -- the full example is available on
+GitHub (URL above).
 \begin{minted}[gobble=4, frame=single]{YAML}
@@
 Taking the demographic model, parameters file, and observed SFS together, we
-can perform inference using the \texttt{moments.Demes.Inference.optimize(...)}
+can perform inference using the \texttt{moments.Demes.Inference.optimize()}
 function. The initial parameter guesses are given by the input demographic
@@
 Confidence intervals may also be computed using
-\texttt{moments.Demes.Inference.uncerts(...)}, once a (locally) optimal model 
-has been found. The following will emit a tab-separated file containing
-parameter names, best-fit values and estimated standard errors.
+\texttt{moments.Demes.Inference.uncerts()}, once a (locally) optimal model has
+been found. The following will emit a tab-separated file containing parameter
+names, best-fit values and estimated standard errors.
 \begin{minted}[gobble=4, frame=single]{python}
@@
 \end{minted}
+Uncertainties may be computed using either the Fisher Information Matrix, which
+tends to underestimate confidence intervals due to nonindependence of linked
+sites, or using the Godambe Information Matrix, which uses a set of
+bootstrapped datasets to correct for this nonindependence in the data
+\citep{coffman2016computationally}. Example 1 in the associated GitHub
+repository demonstrates this using a simulated dataset.
 
-\subsection*{Inferring parameters in a simulated split-with-migration model}
+\subsection*{Example 1: Inferring parameters in a simulated split-with-migration model}
+
+\begin{figure}[t!]
+    \includegraphics{../example1/fig1.pdf}
+    \caption{
+        Inference using simulated data. (A) The true demographic model,
+        including an expansion in the ancestral population and symmetric
+        migration between the descendent populations. (B--D) The simulated
+        data, showing both marginal and the joint SFS. (E--G) Residuals
+        between the data SFS (in D) and the predicted SFS from the inferred
+        demographic models in (H--J), resp.  The two misspecified models (H and
+        I) provide worse fits to the data and both overestimate the split time
+        due to disallowing any size change in the ancestral population.
+    }
+    \label{fig:im}
+\end{figure}
 
@@
 deeper in the past than the true split time. This effect, as demonstrated by
-\citet{momigliano2020biases}, is due to the decreased coalescence rate within
+\citet{momigliano2021biases}, is due to the decreased coalescence rate within
 the ancestral population between the time of the expansion and divergence of
@@
 
-\subsection*{Inferring human-Neanderthal demographic parameters}
+\subsection*{Example 2: Inferring human-Neanderthal demographic parameters}
+
+This example uses publicly available data to infer a demographic model for two
+human groups (a northern European and western African population) and a
+Neanderthal individual. As shown below, even with just a single sample from the
+Neanderthal lineage, we are able to recover key parameters such as the
+human-Neanderthal divergence and introgression proportion.
 
@@
 We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
-were not genotyped in all four existing high-coverage archaic genome sequences, 
-thatlacked high-confidence ancestral state assignments, or fell within 10 kb of 
-exonic or promoter regions and were left with approximately 960 Mb of 
+were not genotyped in the high-coverage archaic genome sequences, 
+that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
+exonic or promoter regions. This left us with approximately 960 Mb of 
 well-characterized and putatively neutrally-evolving sequence.
@@
 
+\begin{figure}[t!]
+    \includegraphics{../example2/fig2.pdf}
+    \caption{
+        Inference using human and Neanderthal data.
+        (A) The best-fit model including the Vindija Neanderthal and two human
+        populations.
+        (B, C) Marginal SFS fits between the model and empirical data are shown
+        for MSL and GBR.
+        (D--I) Marginal two-way spectra predicted by the model and their residuals 
+        against the data are plotted for the three pairs of populations MSL, GBR 
+        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
+    }
+    \label{fig:humans}
+\end{figure}
+
 The model that we wished to fit is relatively simple but still has many 
@@
 first fitting simpler one-- and two--population models.
-Throughout, we used $u=1.5\cdot 10^{-8}$ as an estimate of the mutation rate per 
-generation and 29 years as the average generation time. We began by fitting 
+Throughout, we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per 
+generation and 29 years as the average generation time. When the goal is to infer
+a complex demographic history involving multiple populations, we typically start
+simple, with subsets of the full data, and sequentially add complexity.
+Here, we began by fitting 
 one-population models with ancestral and recent size changes for MSL and GBR.
@@
 We next stitched these one-population models together and refitted the 
-resulting graph to get a two-population model that included continuous symmetric 
+resulting demographic model to get a two-population model 
+that included continuous symmetric 
 migration between MSL and GBR and an ancestral size expansion. Following this,
@@
 Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
-(90 kya, \cite{prufer2017high}) in a second round of optimization.
+\citep[90 kya,][]{prufer2017high} in a second round of optimization.
 Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
 Fixing parameters for which good estimates are available in the literature 
-(see e.g. \cite{sumer2025earliest}) may be reasonable in the first stages of 
+\citep[see e.g.,][]{sumer2025earliest} may be reasonable in the first stages of 
 model optimization, and we expected this value to be especially difficult to fit 
@@
 MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
-percent. We observe some large residuals in our best-fit model 
-Figure~\ref{fig:humans}G-I), suggesting ...
-
-\begin{figure}[t!]
-    \includegraphics{../example2/fig2.pdf}
-    \caption{
-      A plot of the best-fit model created using \demesdraw is displayed in (A).
-      Marginal SFS fits between the model and empirical data are shown for MSL 
-      (B) and GBR (C).
-      Marginal two-way spectra predicted by the model and their residuals 
-      against the data are plotted for the three pairs of populations MSL, GBR 
-      (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
-    }
-    \label{fig:humans}
-\end{figure}
+percent. We observe some large, systematic residuals in our best-fit model 
+(Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
+improved through further exploration of model space.
 
@@
 
-\begin{enumerate}
-    \item Include here general ideas about the strengths and weaknesses of
-        various approaches in population genetic inference, including when
-        using the SFS.
-    \item Challenges in finding local/global optima.
-    \item Challenges in exploring parameter space.
-    \item Background selection [Ewing, Johri]
-    \item Gene dense vs gene sparse genomic architectures among species.
-\end{enumerate}
+The site frequency spectrum is just one of many options for inferring
+evolutionary parameters from genetic data. It is straightforward to compute
+from data, and there are many methods for predicting the SFS, making it an
+appealing framework for demographic inference. As shown here in two examples,
+we have incorporated the \demes format for specifying demographic models in
+\moments to easily compute the SFS and perform demographic inference, including
+computing uncertainties.
+
+\moments is limited in the size of the dataset that it can consider. Sample
+sizes can be large (in the hundreds), especially when considering models with
+one or few populations, but will encounter memory issues when sample sizes are
+extreme or as the number of populations grows. We are limited to a maximum of
+five populations existing at any one time, including ancient samples. Other
+SFS-based methods, such as \momi \citep{kamm2020efficiently, dilber2024faster},
+can include additional populations, typically with smaller sample sizes.
+
+When inferring models with multiple populations, especially four or more, both
+the numbers of parameters and the space of plausible model topologies can grow
+rapidly. It can therefore become challenging for the optimization routines to
+find optima in parameter space, and overparameterization can become a problem.
+One hint that models are overparameterized (or models misspecified) is runaway
+behavior of parameter values, in which they converge to unreasonably small or
+large values (or approach the imposed bounds). Starting with simple
+parameterizations and sequentially adding complexity helps to alleviate this
+issue. The challenge of large model space is more difficult to address, and
+must be addressed through building and testing a wide range of plausible model
+topologies. While time-consuming, a broad exploration of model space,
+especially when considering many populations, is crucial for evaluating
+robustness of particular features of inferred history.
+
+Finally, natural selection distorts allele frequencies not only at sites
+directly under selection, but also in linked regions of the genome through
+background selection. This can have a sizeable impact on the SFS and lead to
+biases in demographic inference \citep{ewing2016consequences, johri2021impact}.
+In gene-sparse species, such as primates, considering data from regions away
+from functionally constrained elements is feasible, as we did in example 2
+here.  Alternatively, the genome may be partitioned by $B$-value \citep[a
+measure of the local strength of background
+selectionk,][]{mcvicker2009widespread} and regions experiencing low levels of
+background selection may be used \citep[for an example,
+see][]{medina2023demographic}. In some species with gene-dense genomes, such as
+\emph{Drosophila} species, this may be infeasible, and interspersed putatively
+neutral variation must be used, such synonymous variation or short introns,
+which will still experience background selection. The appropriate strategy will
+depend on the genome architecture of the species and genomic resources
+available.
+
+\section*{Online resources} \moments is available through \texttt{pypi} as
+\texttt{moments-popgen}, and at \url{https://github.com/MomentsLD/moments} with
+documentation at \url{https://momentsld.github.io/moments/}. The detailed
+online examples accompanying this chapter can be found at
+\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments/}.
 

### 2025-04-01 chapter/paper.tex
minor revisions to chapter

@@
 joint SFS. The first example simulates data with known demographic parameters,
-which we try to reinfer using the both the original model and simpler
+which we try to reinfer using both the original model and simpler
 misspecified models. In the second example, we infer a relatively simple
@@
 consided of 500 regions, each of length 1 Mb, with per-base recombination and
-mutation rates of $10^{-8}$. The joint SFS was computed using \tskit's
+mutation rates of $10^{-8}$. The joint SFS was computed by using \tskit's
 \texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
@@
 we added a Neandertal branch and fitted the 
-Neandertal-modern human divergence time, Neandertal effective size, and the 
+Neandertal-modern human divergence time, the Neandertal effective size, and the 
 Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
@@
 Fixing parameters for which good estimates are available in the literature 
-\citep[see e.g.,][]{sumer2025earliest} may be reasonable in the first stages of 
-model optimization, and we expected this value to be especially difficult to fit 
-due to the small pulse proportion and Neandertal sample size. In the second 
+\citep[see e.g.,][]{sumer2025earliest} is often reasonable in the first stages 
+of model optimization, and we expected this value to be especially difficult to 
+fit due to the small pulse proportion and Neandertal sample size. In the second 
 round of optimization on the three-population model, we refit all parameters 
@@
 appealing framework for demographic inference. As shown here in two examples,
-we have incorporated the \demes format for specifying demographic models in
+we have incorporated the \demes format for specifying demographic models into
 \moments to easily compute the SFS and perform demographic inference, including
@@
 sizes can be large (in the hundreds), especially when considering models with
-one or few populations, but will encounter memory issues when sample sizes are
+one or few populations, but memory issues will develop when sample sizes are
 extreme or as the number of populations grows. We are limited to a maximum of
@@
 issue. The challenge of large model space is more difficult to address, and
-must be addressed through building and testing a wide range of plausible model
+must be handled through building and testing a wide range of plausible model
 topologies. While time-consuming, a broad exploration of model space,
@@
 background selection may be used \citep[for an example,
-see][]{medina2023demographic}. In some species with gene-dense genomes, such as
+see][]{medina2023demographic}. In some species with gene-dense genomes, like
 \emph{Drosophila} species, this may be infeasible, and interspersed putatively
-neutral variation must be used, such synonymous variation or short introns,
+neutral variation must be used, such as synonymous variation or short introns,
 which will still experience background selection. The appropriate strategy will

### 2025-07-31 chapter/paper.tex
start making headway on revisions

--- 2025-04-01 chapter/paper.tex
@@
 \documentclass[]{article}
-\usepackage[round]{natbib}
+\usepackage[numbers,sort,compress]{natbib}
 
-\usepackage{fullpage}
 \usepackage{url}
-\usepackage{authblk}
 \usepackage{graphicx}
@@
 \usepackage{setspace}
+\usepackage{lineno}
 
-% need to have the pythonhighlight.sty in directory
-%\usepackage{pythonhighlight}
+\usepackage{authblk}
+\title{Demographic inference using the SFS with \moments and \demes}
+\author[1]{Nick Collier}
+\author[1,*]{Aaron P. Ragsdale}
+\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison}
+\affil[*]{apragsdale@wisc.edu}
+\affil[ ]{ORCID IDs: 0009-0005-0385-9798 (NC), 0000-0003-0715-3432 (APR)}
+
+\usepackage[margin=1in, headheight=15pt, headsep=10pt]{geometry}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\lhead{N Collier and AP Ragsdale}
+\rhead{Demographic inference with \moments and \demes}
 
@@
 % local definitions
+\usepackage{xspace}
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
-\usepackage{xspace}
+\newcommand{\reviewercomment}[1]{{\textcolor{red}{REVIEWER: #1}}}
 \newcommand{\moments}{\texttt{moments}\xspace}
@@
 \begin{document}
+\linenumbers
+\doublespacing
 
-\title{Demographic inference using the SFS with \moments and \demes}
-\author[1]{Nick Collier}
-%\author[2]{Simon Gravel(?)}
-\author[1,*]{Aaron P. Ragsdale}
-\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison}
-%\affil[2]{McGill University}
-\affil[*]{apragsdale@wisc.edu}
 \maketitle
-
-%\onehalfspacing
 
@@
 
+\emph{Keywords:} Demographic inference, site frequency spectrum, \aprcomment{any others?}
+
+% TODO:
+% - orcid numbers if any - done
+% - interline double, also for references - done
+% - line numbers - done
+% - keywords - done
+% - running head - done
+% - Figures in separate files (named Figure 1, Figure2, etc) for the final submission, legends listed at the end of the chapter. - done
+
 \section*{Introduction}
@@
 chromosome due to recombination (that is, by information stored in the
-Ancestral Recombination Graph \citep{nielsen2025inference}). However, the ARG
-can be large and unwieldy, and methods for reconstructing history directly from
-the ARG, while showing promise \citep[e.g.,][]{yc2022evaluation,
-fan2023likelihood, brandt2024promise}, are in their infancy and so far limited
-in application and scalability. Instead, evolutionary inference using
-informative summaries of genetic variation remains a tractable and powerful
-alternative for learning parameters of population history, natural selection
-and genome biology.
+Ancestral Recombination Graph; see chapters 1, 8 and 11, and
+\cite{nielsen2025inference}). However, the ARG can be large and unwieldy, and
+methods for reconstructing history directly from the ARG, while showing promise
+(e.g., \cite{yc2022evaluation, fan2023likelihood, brandt2024promise}), are in
+their infancy and so far limited in application and scalability. Instead,
+evolutionary inference using informative summaries of genetic variation remains
+a tractable and powerful alternative for learning parameters of population
+history, natural selection and genome biology.
 
@@
 This chapter focuses on multi-population demographic inference from the SFS
-using \moments \citep{jouganous2017inferring}. Past population processes,
-including changes in population size, splits, gene flow and population
-structure, impact the SFS in predictable ways. A population size expansion (or
-contraction) leads to a relative excess (or deficit) of low-frequency variants,
-for example, and commonly used statistics measuring population divergence, such
-as $F_{ST}$, are themselves summaries of the SFS. Thus, the development of
-methods to learn demographic history from the SFS came soon after the
-publication of whole-genome sequencing from multiple individuals
-\citep{marth2004allele, williamson2005simultaneous}. More recent methodological
+using \moments \cite{jouganous2017inferring}.
+
+Briefly, \moments \ldots.
+
+\reviewercomment{A short paragraph
+    briefly describing the principle of the moments methods would be great. How
+    does it differ from other sfs based methods? Just for the readers to get an
+idea, then directing them to the relevent literature for more details.}
+
+Past population processes, including changes in population size, splits, gene flow
+and population structure, impact the SFS in predictable ways. A population size
+expansion (or contraction) leads to a relative excess (or deficit) of
+low-frequency variants, for example, and commonly used statistics measuring
+population divergence, such as $F_{ST}$, are themselves summaries of the SFS.
+Thus, the development of methods to learn demographic history from the SFS came
+soon after the publication of whole-genome sequencing from multiple individuals
+\cite{marth2004allele, williamson2005simultaneous}. More recent methodological
 advances allow for increased sample sizes and numbers of populations, providing
 a rich ecosystem of software for SFS-based demographic inference
-\citep{gutenkunst2009inferring, excoffier2011fastsimcoal,
+\cite{gutenkunst2009inferring, excoffier2011fastsimcoal,
     gravel2011demographic, jouganous2017inferring, ragsdale2018genomic,
@@
 of fitness effects of new mutations, primarily for nonsynonymous variation in
-coding regions \citep[e.g.,][]{eyre2006distribution, boyko2008assessing,
-kim2017inference}; to scan for historical selective sweeps
-\citep[e.g.,][]{kim2002detecting, nielsen2005genomic}; and to understand how
+coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
+kim2017inference}); to scan for historical selective sweeps
+(e.g., \cite{kim2002detecting, nielsen2005genomic}); and to understand how
 selection on quantitative traits impacts the genetic architecture underlying
-those traits \citep[eg.,][]{patel2024conditional, ragsdale2024archaic}. The SFS
+those traits (e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS
 has also been used to infer relative mutation rates and how they have changed
-over time \citep{dewitt2021nonparametric}. Each of these analyses typically
+over time \cite{dewitt2021nonparametric}. Each of these analyses typically
 requires partitioning the data in some way -- by functional annotation,
@@
 made across SFS from different classes of mutations.
+
+\subsection*{Obtaining and installing the software}
+
+\moments is available through \texttt{pypi} as \texttt{moments-popgen} (using
+\texttt{pip install moments-popgen}). Source code can be downloaded from
+\url{https://github.com/MomentsLD/moments}, and  documentation is hosted at
+\url{https://momentsld.github.io/moments/}. The detailed online examples
+accompanying this chapter can be found at
+\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments/}, which uses
+\moments version 1.4 and \demes version 0.2.
 
@@
 their relationships via splits and gene flow. \demes provides a standardized
-and accessible format for defining demographic models \citep{gower2022demes},
+and accessible format for defining demographic models \cite{gower2022demes},
 and has been adopted by widely used software for population genetics simulation
-and inference (including \msprime \citep{baumdicker2022efficient}, \momi
-\citep{dilber2024faster}, \fwdpy \citep{thornton2019polygenic}, and
-\texttt{GADMA} \citep{noskova2023gadma2}). \demes encourages interoperability
+and inference (including \msprime \cite{baumdicker2022efficient}, \momi
+\cite{dilber2024faster}, \fwdpy \cite{thornton2019polygenic}, and
+\texttt{GADMA} \cite{noskova2023gadma2}). \demes encourages interoperability
 and reuse of code across software, ease of model implementation and reduction
@@
 
-Demographic models are specified in YAML format (see \citet{gower2022demes} and
+Demographic models are specified in YAML format (see \cite{gower2022demes} and
 the associated documentation). To illustrate, the two-population
@@
 below:
+
+\singlespacing
 \begin{minted}[gobble=4, frame=single,]{yaml}
@@
 \end{minted}
+\doublespacing
 
 In \moments, a \demes-specified demographic model can be directly used to
-compute either the SFS or multi-population LD statistics
-\citep{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
+compute expectations for either the SFS or multi-population LD statistics
+\cite{ragsdale2019models, ragsdale2020unbiased}. For the SFS, we simply need
 to load the demographic model using \demes and then specify the number of
@@
 
+\reviewercomment{Would it be possible to include some information regarding possible local optima? Do you recommend starting from different initial conditions?}
+
 In order to optimize parameters in a \demes model, we need a way to specify
@@
 GitHub (URL above).
+
+\singlespacing
 \begin{minted}[gobble=4, frame=single]{YAML}
@@
 \end{minted}
+\doublespacing
 
@@
 we have an estimate for the total mutation rate, $U$.
+
+\singlespacing
 \begin{minted}[gobble=4, frame=single]{python}
@@
 names, best-fit values and estimated standard errors.
+
+\singlespacing
 \begin{minted}[gobble=4, frame=single]{python}
@@
 \end{minted}
+\doublespacing
+
 Uncertainties may be computed using either the Fisher Information Matrix, which
@@
 bootstrapped datasets to correct for this nonindependence in the data
-\citep{coffman2016computationally}. Example 1 in the associated GitHub
+\cite{coffman2016computationally}. Example 1 in the associated GitHub
 repository demonstrates this using a simulated dataset.
@@
 
-\begin{figure}[t!]
-    \includegraphics{../example1/fig1.pdf}
-    \caption{
-        Inference using simulated data. (A) The true demographic model,
-        including an expansion in the ancestral population and symmetric
-        migration between the descendent populations. (B--D) The simulated
-        data, showing both marginal and the joint SFS. (E--G) Residuals
-        between the data SFS (in D) and the predicted SFS from the inferred
-        demographic models in (H--J), resp.  The two misspecified models (H and
-        I) provide worse fits to the data and both overestimate the split time
-        due to disallowing any size change in the ancestral population.
-    }
-    \label{fig:im}
-\end{figure}
-
 Using the split-with-migration model defined above (illustrated in
 Figure~\ref{fig:im}A), we simulated data for 30 diploid individuals from both
-populations using \msprime \citep{baumdicker2022efficient}. These simulations
+populations using \msprime \cite{baumdicker2022efficient}. These simulations
 consided of 500 regions, each of length 1 Mb, with per-base recombination and
@@
 the ancestral population, this split time is inferred to be substantially
-deeper in the past than the true split time. This effect, as demonstrated by
-\citet{momigliano2021biases}, is due to the decreased coalescence rate within
+deeper in the past than the true split time. This effect, as demonstrated in
+\cite{momigliano2021biases}, is due to the decreased coalescence rate within
 the ancestral population between the time of the expansion and divergence of
@@
 We obtained modern human genome sequences from a recent resequencing of the 
-1000 Genomes Project cohort \citep{byrska2022high}, and used the high-coverage 
+1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
 sequence of the Vindija Neandertal from \cite{prufer2017high}. 
@@
 
-\begin{figure}[t!]
-    \includegraphics{../example2/fig2.pdf}
-    \caption{
-        Inference using human and Neanderthal data.
-        (A) The best-fit model including the Vindija Neanderthal and two human
-        populations.
-        (B, C) Marginal SFS fits between the model and empirical data are shown
-        for MSL and GBR.
-        (D--I) Marginal two-way spectra predicted by the model and their residuals 
-        against the data are plotted for the three pairs of populations MSL, GBR 
-        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
-    }
-    \label{fig:humans}
-\end{figure}
-
 The model that we wished to fit is relatively simple but still has many 
-parameters (11 in the example we present), so we constructed it in stages by 
+parameters ($11$ in the example we present), so we constructed it in stages by 
 first fitting simpler one-- and two--population models.
@@
 
+\reviewercomment{maybe it would be helpful to also include the likelihood (or AIC?) of each step to show how it changes with the addition of further parameters, and how it can be used to inform the models design}
+\aprcomment{this suggestion would only make sense for models fit to the same data SFS, but it could be done to compare the fit without vs.
+with the Neanderthal pulse.
+Maybe we could demonstrate the likelihood ratio test, if it works?
+what do you think, Nick?}
+
 We next stitched these one-population models together and refitted the 
@@
 Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
-\citep[90 kya,][]{prufer2017high} in a second round of optimization.
+(90 kya, \cite{prufer2017high}) in a second round of optimization.
 Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
 Fixing parameters for which good estimates are available in the literature 
-\citep[see e.g.,][]{sumer2025earliest} may be reasonable in the first stages of 
+(see, e.g., \cite{sumer2025earliest}) may be reasonable in the first stages of 
 model optimization, and we expected this value to be especially difficult to fit 
@@
 five populations existing at any one time, including ancient samples. Other
-SFS-based methods, such as \momi \citep{kamm2020efficiently, dilber2024faster},
+SFS-based methods, such as \momi \cite{kamm2020efficiently, dilber2024faster},
 can include additional populations, typically with smaller sample sizes.
+
+\reviewercomment{Mirroring my comments in the introduction, I would suggest to add a few sentences to highlight how moments differ from other approaches, in terms of caveats and advantages.
+No need to go in details, just the take-home message and references for further details.}
 
@@
 background selection. This can have a sizeable impact on the SFS and lead to
-biases in demographic inference \citep{ewing2016consequences, johri2021impact}.
+biases in demographic inference \cite{ewing2016consequences, johri2021impact}.
 In gene-sparse species, such as primates, considering data from regions away
 from functionally constrained elements is feasible, as we did in example 2
-here.  Alternatively, the genome may be partitioned by $B$-value \citep[a
+here.  Alternatively, the genome may be partitioned by $B$-value (a
 measure of the local strength of background
-selectionk,][]{mcvicker2009widespread} and regions experiencing low levels of
-background selection may be used \citep[for an example,
-see][]{medina2023demographic}. In some species with gene-dense genomes, such as
+selection, \cite{mcvicker2009widespread}) and regions experiencing low levels of
+background selection may be used (for an example,
+see \cite{medina2023demographic}). In some species with gene-dense genomes, such as
 \emph{Drosophila} species, this may be infeasible, and interspersed putatively
@@
 
-\section*{Online resources} \moments is available through \texttt{pypi} as
-\texttt{moments-popgen}, and at \url{https://github.com/MomentsLD/moments} with
-documentation at \url{https://momentsld.github.io/moments/}. The detailed
-online examples accompanying this chapter can be found at
-\url{https://github.com/StatisticalPopulationGenomics-2ndEd/moments/}.
-
-\bibliographystyle{plainnat}
+\bibliographystyle{unsrt}
 \bibliography{paper}
 
+\clearpage
+
+\begin{figure}
+    %\includegraphics{../example1/fig1.pdf}
+    \caption{
+        Inference using simulated data. (A) The true demographic model,
+        including an expansion in the ancestral population and symmetric
+        migration between the descendent populations. (B--D) The simulated
+        data, showing both marginal and the joint SFS. (E--G) Residuals
+        between the data SFS (in D) and the predicted SFS from the inferred
+        demographic models in (H--J), resp.  The two misspecified models (H and
+        I) provide worse fits to the data and both overestimate the split time
+        due to disallowing any size change in the ancestral population.
+        \reviewercomment{it might be good to label the colorbars for the joint spectra as 'allele frequency' for people that are unfamiliar with what those plots are showing}
+    }
+    \label{fig:im}
+\end{figure}
+
+\begin{figure}
+    %\includegraphics{../example2/fig2.pdf}
+    \caption{
+        Inference using human and Neanderthal data.
+        (A) The best-fit model including the Vindija Neanderthal and two human
+        populations.
+        (B, C) Marginal SFS fits between the model and empirical data are shown
+        for MSL and GBR.
+        (D--I) Marginal two-way spectra predicted by the model and their residuals 
+        against the data are plotted for the three pairs of populations MSL, GBR 
+        (D and G), MSL, Vindija (E and H), and GBR, Vindija (F, I).
+    }
+    \label{fig:humans}
+\end{figure}
+
 \end{document}

### 2025-07-31 chapter/paper.tex

@@
 parameters ($11$ in the example we present), so we constructed it in stages by
-first fitting simpler one-- and two--population models.
-Throughout, we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per 
-generation and 29 years as the average generation time. When the goal is to infer
-a complex demographic history involving multiple populations, we typically start
-simple, with subsets of the full data, and sequentially add complexity.
-Here, we began by fitting 
-one-population models with ancestral and recent size changes for MSL and GBR.
-For the MSL model, we allowed three epochs with constant effective size in each 
-epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
-exponential growth to represent the out-of-Africa bottleneck and subsequent 
-rapid expansion. YAML files representing the models and parameters in this 
-section can be found in the linked GitHub repository.
+first fitting simpler one- and two-population models.  Throughout, we used
+$u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation and 29
+years as the average generation time. When the goal is to infer a complex
+demographic history involving multiple populations, we typically start simple,
+with subsets of the full data, and sequentially add complexity.  Here, we began
+by fitting one-population models with ancestral and recent size changes for MSL
+and GBR.  For the MSL model, we allowed three epochs with constant effective
+size in each epoch. We attributed to GBR a sharp contraction $\sim$60 kya
+followed by exponential growth to represent the out-of-Africa bottleneck and
+subsequent rapid expansion. YAML files representing the models and parameters
+in this section can be found in the linked GitHub repository.
 
@@
 Fixing parameters for which good estimates are available in the literature 
-(see, e.g., \cite{sumer2025earliest}) may be reasonable in the first stages of 
+(see, e.g., \cite{sumer2025earliest}) is often reasonable in the first stages of 
 model optimization, and we expected this value to be especially difficult to fit 
@@
 from functionally constrained elements is feasible, as we did in example 2
-here.  Alternatively, the genome may be partitioned by $B$-value (a
-measure of the local strength of background
-selection, \cite{mcvicker2009widespread}) and regions experiencing low levels of
-background selection may be used (for an example,
-see \cite{medina2023demographic}). In some species with gene-dense genomes, such as
-\emph{Drosophila} species, this may be infeasible, and interspersed putatively
-neutral variation must be used, such synonymous variation or short introns,
-which will still experience background selection. The appropriate strategy will
-depend on the genome architecture of the species and genomic resources
-available.
+here.  Alternatively, the genome may be partitioned by $B$-value (a measure of
+the local strength of background selection, \cite{mcvicker2009widespread}) and
+regions experiencing low levels of background selection may be used (for an
+example, see \cite{medina2023demographic}). In some species with gene-dense
+genomes, like \emph{Drosophila} species, this may be infeasible, and
+interspersed putatively neutral variation must be used, such as synonymous
+variation or short introns, which will still experience background selection.
+The appropriate strategy will depend on the genome architecture of the species
+and genomic resources available.
 

### 2025-08-25 chapter/paper.tex
Added a null model, better SFS parsing scripts and edited manuscript

@@
 \title{Demographic inference using the SFS with \moments and \demes}
-\author[1]{Nick Collier}
+\author[1]{Nicholas W. Collier}
 \author[1,*]{Aaron P. Ragsdale}
@@
 \affil[*]{apragsdale@wisc.edu}
-\affil[ ]{ORCID IDs: 0009-0005-0385-9798 (NC), 0000-0003-0715-3432 (APR)}
+\affil[ ]{ORCID IDs: 0009-0005-0385-9798 (NWC), 0000-0003-0715-3432 (APR)}
 
@@
 \pagestyle{fancy}
-\lhead{N Collier and AP Ragsdale}
+\lhead{NW Collier and AP Ragsdale}
 \rhead{Demographic inference with \moments and \demes}
@@
 misspecified models. In the second example, we infer a relatively simple
-human-Neanderthal history using data from two human populations and a
-high-coverage ancient sample from the Neanderthal lineage. For both of these
+human-Neandertal history using data from two human populations and a
+high-coverage ancient sample from the Neandertal lineage. For both of these
 examples, we briefly describe data processing and inference setup, emphasizing
@@
 
-\subsection*{Example 2: Inferring human-Neanderthal demographic parameters}
+\subsection*{Example 2: Inferring human-Neandertal demographic parameters}
 
 This example uses publicly available data to infer a demographic model for two
-human groups (a northern European and western African population) and a
-Neanderthal individual. As shown below, even with just a single sample from the
-Neanderthal lineage, we are able to recover key parameters such as the
-human-Neanderthal divergence and introgression proportion.
+modern human populations (a northern European and western African population) 
+and a Neandertal individual. As shown below, even with just a single sample 
+from the Neandertal lineage, we are able to recover key parameters such as the
+human-Neandertal divergence time and fraction of ancestry contributed to modern 
+Europeans by Neandertals.
 
@@
 1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
-sequence of the Vindija Neandertal from \cite{prufer2017high}. 
+sequence of the Vindija Neandertal genome published by \cite{prufer2017high}. 
 From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
@@
 parameters ($11$ in the example we present), so we constructed it in stages by
-first fitting simpler one- and two-population models.  Throughout, we used
-$u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation and 29
-years as the average generation time. When the goal is to infer a complex
-demographic history involving multiple populations, we typically start simple,
-with subsets of the full data, and sequentially add complexity.  Here, we began
-by fitting one-population models with ancestral and recent size changes for MSL
-and GBR.  For the MSL model, we allowed three epochs with constant effective
-size in each epoch. We attributed to GBR a sharp contraction $\sim$60 kya
-followed by exponential growth to represent the out-of-Africa bottleneck and
-subsequent rapid expansion. YAML files representing the models and parameters
-in this section can be found in the linked GitHub repository.
+first fitting marginal one- and two-population models. When our goal is to infer 
+a complex demographic history involving multiple populations, it is helpful to 
+start simple, with subsets of the full data, and sequentially add complexity.  
+Here, we began by fitting one-population models with ancestral and recent size 
+changes for the MSL and GBR populations.  For the MSL model, we allowed three 
+epochs with constant effective size in each epoch. We attributed to GBR a sharp 
+contraction $\sim$60 kya followed by exponential growth to represent the 
+out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
+we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
+and 29 years as the average generation time. YAML files representing the models 
+and parameters in this section can be found in the linked GitHub repository.
 
-\reviewercomment{maybe it would be helpful to also include the likelihood (or AIC?) of each step to show how it changes with the addition of further parameters, and how it can be used to inform the models design}
-\aprcomment{this suggestion would only make sense for models fit to the same data SFS, but it could be done to compare the fit without vs.
-with the Neanderthal pulse.
-Maybe we could demonstrate the likelihood ratio test, if it works?
-what do you think, Nick?}
-
-We next stitched these one-population models together and refitted the 
-resulting demographic model to get a two-population model 
-that included continuous symmetric 
-migration between MSL and GBR and an ancestral size expansion. Following this,
-we added a Neandertal branch and fitted the 
+We next stitched these one-population models together and refitted the resulting 
+demographic model to get a two-population model that included continuous 
+symmetric migration between MSL and GBR and an ancestral size expansion. 
+Following this, we added a Neandertal branch and fitted the 
 Neandertal-modern human divergence time, the Neandertal effective size, and the 
-Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
-between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
-admixed with modern non-African humans. Finding that this parameter tended to 
-converge to its lower bound (imposed by the estimated time of death of the 
-Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
-(90 kya, \cite{prufer2017high}) in a second round of optimization.
-Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
-Fixing parameters for which good estimates are available in the literature 
-(see, e.g., \cite{sumer2025earliest}) is often reasonable in the first stages of 
-model optimization, and we expected this value to be especially difficult to fit 
-due to the small pulse proportion and Neandertal sample size. In the second 
-round of optimization on the three-population model, we refit all parameters 
-aside from the two event times mentioned above to refine the estimates.
+Neandertal-GBR pulse proportion, leaving other parameters fixed. We also 
+fitted the divergence time between the sampled Vindija Neandertal lineage and 
+the Neandertal lineage which admixed with modern non-African humans. Finding 
+that this parameter tended to converge to its lower bound (imposed by the 
+estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
+value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
+of optimization. Rather than fitting it, we fixed the time of the Neandertal-GBR 
+pulse to 50 kya. Fixing parameters for which good estimates are available in the 
+literature  (see, e.g., \cite{sumer2025earliest}) is often reasonable in the 
+first stages of model optimization, and we expected this value to be especially 
+poorly constrained due to the small pulse proportion and Neandertal sample size. 
+In the second round of optimization on the three-population model, we refit all 
+parameters aside from the two event times mentioned above to refine the 
+estimates.
 
 After this refitting step, we were left with parameters in good agreement with 
-prior authors, with a Neandertal-modern human divergence time of 547 kya, 
-MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion of 2.7 
-percent. We observe some large, systematic residuals in our best-fit model 
+those inferred by prior authors, with a Neandertal-modern human divergence time 
+of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
+of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
 (Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
 improved through further exploration of model space.
+
+We also fitted a nested model without the Neanderthal-GBR pulse, which had a 
+substantially lower likelihood than the pulse model (log-likelihood [LL] 
+$\approx -127,552$, against LL $\approx -25,103$ for the pulse model). We did 
+not perform a formal hypothesis test; the null distribution of the 
+likelihood-ratio test (LRT) statistic is overdispersed under composite 
+likelihood, inflating the rate of false positives and rendering its use 
+unjustifiable in this setting. \moments does incorporate a function 
+(\texttt{moments.Godambe.LRT\_adjust}) which computes a multiplicative
+adjustment, intended to restore the expected null distribution, using 
+first-order moment matching \cite{coffman2016computationally}. This method 
+requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
+test by fitting the nested and pulse models to a large number of bootstrap 
+samples and calculating the rate of rejection across replicates-- which has the 
+benefit of producing a distribution of LRT statistics, but is time-consuming
+and computationally expensive.
 
@@
     \caption{
-        Inference using human and Neanderthal data.
-        (A) The best-fit model including the Vindija Neanderthal and two human
+        Inference using human and Neandertal data.
+        (A) The best-fit model including the Vindija Neandertal and two human
         populations.

### 2025-08-25 example2/data/bootstrap_fs.py

@@ -1,0 +1,42 @@
+# bootstrap over site frequency spectra. also take their sum.
+
+import numpy as np
+import moments 
+import pickle
+import pandas
+import random
+import os 
+
+
+fnames = sorted(os.listdir("spectra/"))
+region_fs = dict()
+for fname in fnames: 
+    region = ".".join(fname.split(".")[2:4])
+    region_fs[region] = moments.Spectrum.from_file("spectra/" + fname)
+
+
+L_tbl = pandas.read_csv("region_L_tbl.csv")
+L_tot = np.sum(L_tbl["L"])
+
+
+# resample bootstrap replicates.
+n_reps = 100
+n_samps = len(region_fs)
+boot_reps = []
+for ii in range(n_reps):
+    draws = random.choices(list(region_fs.keys()), k=n_samps)
+    boot_fs = 0
+    boot_L = 0
+    for region in draws:
+        boot_fs += region_fs[region]
+        boot_L += np.array(L_tbl[L_tbl["region"] == region]["L"])[0]
+    boot_reps.append((boot_fs, boot_L))
+with open("bootstrap_data.pkl", "wb") as fout:
+    pickle.dump(boot_reps, fout)
+
+
+# take the sum
+sum_fs = 0
+for region in region_fs:
+    sum_fs += region_fs[region]
+sum_fs.to_file("MSL_GBR_Vindija.fs")

### 2025-08-25 example2/data/compute_L.py

@@ -1,0 +1,43 @@
+# compute L for each window and save the result
+
+import numpy as np
+import moments 
+import pandas
+
+
+bed_file = lambda chrom: f"bed_files/combined_mask_chr{chrom}.bed.gz"
+
+
+# interval size
+l = 30000000
+
+
+def load_genome_file(fname):
+    ret = dict()
+    with open(fname, "r") as fin:
+        for line in fin:
+            chrom, L = line.split()
+            ret[chrom] = int(L)
+    return ret 
+
+
+genome_file = "hg19.genome"
+seq_lens = load_genome_file(genome_file)
+
+
+# we loop over chromosomes and intervals
+Ls = {"region": [], "interval": [], "L": []}
+for chrom in range(1, 23):
+    length = seq_lens[f"chr{chrom}"]
+    intervals = [[x, y] for x, y 
+                 in zip(range(1, length, l), range(l + 1, length + l, l))]
+    for ii, interval in enumerate(intervals):
+        L = moments.Parsing.compute_L(bed_file(chrom), interval=interval)
+        Ls["region"].append(f"chrom_{chrom}.region_{ii}")
+        Ls["interval"].append(interval)
+        Ls["L"].append(L)
+
+
+df = pandas.DataFrame(Ls)
+df.to_csv("region_L_tbl.csv", index=False)
+

### 2025-08-25 example2/data/parse_fs.py

@@ -1,0 +1,48 @@
+# compute the frequency spectra across all chromosomes and save them
+
+import numpy as np
+import moments 
+
+
+pop_file = "populations.txt"
+vcf_file = lambda chrom: f"vcf_files/annotated_variants_chr{chrom}.vcf.gz"
+bed_file = lambda chrom: f"bed_files/combined_mask_chr{chrom}.bed.gz"
+
+
+# interval size
+l = 30000000
+
+
+def load_genome_file(fname):
+    ret = dict()
+    with open(fname, "r") as fin:
+        for line in fin:
+            chrom, L = line.split()
+            ret[chrom] = int(L)
+    return ret 
+
+
+genome_file = "hg19.genome"
+seq_lens = load_genome_file(genome_file)
+
+
+# we loop over chromosomes and intervals
+for chrom in range(1, 23):
+    L = seq_lens[f"chr{chrom}"]
+    intervals = [[x, y] for x, y in zip(range(1, L, l), range(l + 1, L + l, l))]
+    for ii, interval in enumerate(intervals):
+        try:
+            fs = moments.Spectrum.from_vcf(
+                vcf_file(chrom),
+                bed_file=bed_file(chrom),
+                pop_file=pop_file,
+                pops=["MSL", "GBR", "Vindija"],
+                use_AA=True,
+                verbose=100000,
+                interval=interval,
+            )
+            fs_proj = fs.project([20, 20, 2])
+            fs_proj.to_file(f"spectra/3pop.projected.chrom_{chrom}.region_{ii}.fs")
+        except:
+            print(f"Empty interval chr{chrom} {interval}")
+

### 2025-08-25 example2/fit_MSL_GBR_Vindija_model_step1.py

@@ -11,7 +11,7 @@ graph_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_model.yaml"
 options_file = "models/MSL_GBR_Vindija_step1/MSL_GBR_Vindija_options.yaml"
 
 # define data path and load data
-data_file = "spectra/MSL_GBR_Vindija"
+data_file = "spectra/MSL_GBR_Vindija.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename

### 2025-08-25 example2/fit_MSL_GBR_Vindija_model_step2.py

@@ -11,7 +11,7 @@ graph_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_model_step2.yaml"
 options_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_options_step2.yaml"
 
 # define data path and load data
-data_file = "spectra/MSL_GBR_Vindija"
+data_file = "spectra/MSL_GBR_Vindija.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename

### 2025-08-25 example2/fit_MSL_GBR_Vindija_null_model.py

@@ -1,4 +1,4 @@
-## Fit a two-epoch model for GBR
+## Fit parameters to the Neandertal branch of a three-deme MSL, GBR, Vindija model
 ## Then compute confidence intervals and plot the fit
 
 import demes, demesdraw
@@ -7,15 +7,15 @@ import matplotlib.pyplot as plt
 
 
 # define model paths
-graph_file = "models/GBR/GBR_model.yaml"
-options_file = "models/GBR/GBR_options.yaml"
+graph_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.yaml"
+options_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model_options.yaml"
 
 # define data path and load data
-data_file = "spectra/GBR.fs"
+data_file = "spectra/MSL_GBR_Vindija.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename
-output_file = "models/GBR/GBR_model.misid_fit.yaml"
+output_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_fit.yaml"
 
 # define parameters
 u = 1.5e-8
@@ -32,7 +32,7 @@ param_names, fit_params, ll = moments.Demes.Inference.optimize(
     fit_ancestral_misid=True, 
     misid_guess=misid_guess,
     uL=U, 
-    verbose=50,
+    verbose=1,
     output=output_file, 
     method="powell",
     overwrite=True
@@ -48,7 +48,7 @@ for name, value in zip(param_names, fit_params):
 p_misid = fit_params[-1]
 
 # compute confidence intervals using FIM 
-log_file = "models/GBR/GBR_model.misid_fit.uncerts.txt"
+log_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_fit.uncerts.txt" 
 uncerts = moments.Demes.Inference.uncerts(
     output_file,
     options_file,
@@ -60,7 +60,7 @@ uncerts = moments.Demes.Inference.uncerts(
     verbose=10,
     output=log_file,
     overwrite=True,
-    eps=1e-3
+    eps=1e-2
 )
 
 # print confidence intervals
@@ -73,17 +73,31 @@ for name, value, err in zip(param_names, fit_params, uncerts):
[…]

### 2025-08-25 example2/fit_MSL_GBR_model.py

@@ -11,7 +11,7 @@ graph_file = "models/MSL_GBR/MSL_GBR_model.yaml"
 options_file = "models/MSL_GBR/MSL_GBR_options.yaml"
 
 # define data path and load data
-data_file = "spectra/MSL_GBR"
+data_file = "spectra/MSL_GBR.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename

### 2025-08-25 example2/fit_MSL_model.py

@@ -11,7 +11,7 @@ graph_file = "models/MSL/MSL_model.yaml"
 options_file = "models/MSL/MSL_options.yaml"
 
 # define data path and load data
-data_file = "spectra/MSL"
+data_file = "spectra/MSL.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename

### 2025-08-25 example2/perform_lrt.py

@@ -1,0 +1,80 @@
+# Perform the LRT between MSL_GBR_Vindija_null_model and MSL_GBR_Vindija_step2
+
+import demes
+import moments
+import numpy as np
+import pickle
+
+
+# parameters
+u = 1.5e-8
+L = 960914001
+U = u * L
+samples = {"MSL": 20, "GBR": 20, "Vindija": 2}
+
+
+# models
+model_0_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_fit.yaml"
+misid_0 = 0.0184
+model_1_file =  "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_model_step2.misid_fit.yaml"
+misid_1 = 0.0223
+
+
+# data
+data_file = "spectra/MSL_GBR_Vindija.fs"
+data = moments.Spectrum.from_file(data_file)
+
+bootstrap_data_file = "data/bootstrap_data.pkl"
+with open(bootstrap_data_file, "rb") as fin:
+    all_boot = pickle.load(fin)
+all_boot = [x[0] for x in all_boot]
+
+
+graph_0 = demes.load(model_0_file)
+model_0 = moments.Demes.SFS(graph_0, samples=samples, u=u, L=L)
+model_0 = moments.Misc.flip_ancestral_misid(model_0, misid_0)
+ll_0 = moments.Inference.ll(model_0, data)
+
+graph_1 = demes.load(model_1_file)
+model_1 = moments.Demes.SFS(graph_1, samples=samples, u=u, L=L)
+model_1 = moments.Misc.flip_ancestral_misid(model_1, misid_1)
+ll_1 = moments.Inference.ll(model_1, data)
+
+
+# load parameter values from nested model
+options_0_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model_options.yaml"
+options_0 = moments.Demes.Inference._get_params_dict(options_0_file)
+builder_0 = demes.load(model_0_file).asdict()
+pnames, p0, *_ = moments.Demes.Inference._set_up_params_and_bounds(options_0, builder_0)
+
+nested_idx = np.array([3])
+adj_params = np.insert(p0, 3, 0.0)
+
+
+options_1_file = "models/MSL_GBR_Vindija_step2/MSL_GBR_Vindija_options_step2.yaml"
+builder_1 = demes.load_asdict(model_1_file)
[…]

### 2025-08-25 example2/refit_MSL_GBR_Vindija_null_model.py

@@ -1,4 +1,4 @@
-## Fit a two-epoch model for GBR
+## Fit parameters to the Neandertal branch of a three-deme MSL, GBR, Vindija model
 ## Then compute confidence intervals and plot the fit
 
 import demes, demesdraw
@@ -7,21 +7,21 @@ import matplotlib.pyplot as plt
 
 
 # define model paths
-graph_file = "models/GBR/GBR_model.yaml"
-options_file = "models/GBR/GBR_options.yaml"
+graph_file =  "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_fit.yaml"
+options_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model_options.yaml"
 
 # define data path and load data
-data_file = "spectra/GBR.fs"
+data_file = "spectra/MSL_GBR_Vindija.fs"
 data = moments.Spectrum.from_file(data_file)
 
 # define output graph filename
-output_file = "models/GBR/GBR_model.misid_fit.yaml"
+output_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_refit.yaml"
 
 # define parameters
 u = 1.5e-8
 L = 960914001
 U = u * L
-misid_guess = 0.03
+misid_guess = 0.0184
 
 # fit using the Powell algorithm
 param_names, fit_params, ll = moments.Demes.Inference.optimize(
@@ -32,7 +32,7 @@ param_names, fit_params, ll = moments.Demes.Inference.optimize(
     fit_ancestral_misid=True, 
     misid_guess=misid_guess,
     uL=U, 
-    verbose=50,
+    verbose=1,
     output=output_file, 
     method="powell",
     overwrite=True
@@ -48,7 +48,7 @@ for name, value in zip(param_names, fit_params):
 p_misid = fit_params[-1]
 
 # compute confidence intervals using FIM 
-log_file = "models/GBR/GBR_model.misid_fit.uncerts.txt"
+log_file = "models/MSL_GBR_Vindija_null_model/MSL_GBR_Vindija_null_model.misid_refit.uncerts.txt" 
 uncerts = moments.Demes.Inference.uncerts(
     output_file,
     options_file,
@@ -60,7 +60,7 @@ uncerts = moments.Demes.Inference.uncerts(
     verbose=10,
     output=log_file,
     overwrite=True,
-    eps=1e-3
[…]

### 2025-08-25 example2/spectra/marginalize_spectra.py

@@ -1,0 +1,14 @@
+
+import moments 
+
+
+full_fs = moments.Spectrum.from_file("MSL_GBR_Vindija.fs")
+
+fs_MSL = full_fs.marginalize([1, 2])
+fs_MSL.to_file("MSL.fs")
+
+fs_GBR = full_fs.marginalize([0, 2])
+fs_GBR.to_file("GBR.fs")
+
+fs_MSL_GBR = full_fs.marginalize([2])
+fs_MSL_GBR.to_file("MSL_GBR.fs")

### 2025-08-26 chapter/paper.tex
Update text, revise figure 1

@@
 
-\emph{Keywords:} Demographic inference, site frequency spectrum, \aprcomment{any others?}
-
-% TODO:
-% - orcid numbers if any - done
-% - interline double, also for references - done
-% - line numbers - done
-% - keywords - done
-% - running head - done
-% - Figures in separate files (named Figure 1, Figure2, etc) for the final submission, legends listed at the end of the chapter. - done
+\emph{Keywords:} Demographic inference, site frequency spectrum, diffusion
+equation, moments equations
 
@@
 
-This chapter focuses on multi-population demographic inference from the SFS
-using \moments \cite{jouganous2017inferring}.
-
-Briefly, \moments \ldots.
-
-\reviewercomment{A short paragraph
-    briefly describing the principle of the moments methods would be great. How
-    does it differ from other sfs based methods? Just for the readers to get an
-idea, then directing them to the relevent literature for more details.}
-
 Past population processes, including changes in population size, splits, gene flow
@@
 
-Other evolutionary mechanisms are known to affect allele frequency dynamics and
+This chapter focuses on multi-population demographic inference from the SFS
+using \moments \cite{jouganous2017inferring}. Briefly, \moments computes the
+expected SFS by numerically solving a system of ordinary differential equations
+closely related to the diffusion approximation \cite{gutenkunst2009inferring}.
+The flexibility of the numerical methods in \moments allows for non-constant
+population size histories including up to five populations at a time, with gene
+flow. Selection can also be incorporated, allowing for recessive, additive or
+dominant selection, or over- or underdominance at a locus. This differs from
+most coalescent-based approaches (e.g.,
+\cite{excoffier2011fastsimcoal,dewitt2021nonparametric,dilber2024faster}),
+which are typically limited to neutrality.
+
+Many evolutionary mechanisms are known to affect allele frequency dynamics and
 the SFS. This presents a challenge for demographic inference, as the observed
@@
 
-\reviewercomment{Would it be possible to include some information regarding possible local optima? Do you recommend starting from different initial conditions?}
-
 In order to optimize parameters in a \demes model, we need a way to specify
@@
 function. The initial parameter guesses are given by the input demographic
-model, which may be perturbed using a keyword argument. Any value in the
-demographic model that is not specified in the parameters file will remain
-unchanged, and is therefore treated as a fixed parameter. Here, we also assume
-we have an estimate for the total mutation rate, $U$.
+model, which may be perturbed using a keyword argument. The likelihood surface
+may be rugged or flat in some dimensions, and optimization routines sometimes
+fail to converge to the global (or sometimes even local) optimum. Thus, we
+often try many different initial guesses for parameter values, which can be
+randomly assigned using the \texttt{perturb} keyword argument. Running multiple
+iterations of optimization with different optimization functions is also
+recommended to ensure convergence.
+
+Any value in the demographic model that is not specified in the parameters file
+will remain unchanged, and is therefore treated as a fixed parameter. Here, we
+also assume we have an estimate for the total mutation rate, $U$.
 
@@
 
-\reviewercomment{Mirroring my comments in the introduction, I would suggest to add a few sentences to highlight how moments differ from other approaches, in terms of caveats and advantages.
-No need to go in details, just the take-home message and references for further details.}
-
 When inferring models with multiple populations, especially four or more, both
@@
         due to disallowing any size change in the ancestral population.
-        \reviewercomment{it might be good to label the colorbars for the joint spectra as 'allele frequency' for people that are unfamiliar with what those plots are showing}
     }

### 2025-08-26 example1/plot_fig1.py

@@ -127,6 +127,8 @@ fig.text(0.05, 0.48, "H", fontsize=7, va="center", ha="center")
 fig.text(0.35, 0.48, "I", fontsize=7, va="center", ha="center")
 fig.text(0.67, 0.48, "J", fontsize=7, va="center", ha="center")
 
+fig.text(0.75, 0.95, "Count", fontsize=6, va="center", ha="center")
+
 fig.savefig("fig1.pdf")
 #fig.show()
 

### 2025-08-26 chapter/paper.tex
Nicks final revisions to second draft

@@
 \newcommand{\reviewercomment}[1]{{\textcolor{red}{REVIEWER: #1}}}
+\newcommand{\nwccomment}[1]{{\textcolor{magenta}{NWC: #1}}}
 \newcommand{\moments}{\texttt{moments}\xspace}
@@
 misspecified models. In the second example, we infer a relatively simple
-human-Neandertal history using data from two human populations and a
+Neanderthal-human history using data from two human populations and a
 high-coverage ancient sample from the Neandertal lineage. For both of these
@@
 In order to optimize parameters in a \demes model, we need a way to specify
-which parameters should be fit. For this, we use a separate YAML-formatted file
+which parameters should be fitted. For this, we use a separate YAML-formatted file
 to define \texttt{parameters}, each of which points to desired value(s) in the
@@
 
-We fit three demographic models to the simulated data. In addition to
-reinferring the parameters from the simulated model, we fit two simpler models
-to the same data. These had fewer parameters, both omitting the size change
-deeper in time in the ancestral population. This is meant to crudely mimic the
-scenario that we often face, in which the true history is more complicated than
-the parameterized model. It also highlights biases in inferences that can arise
-when features of the true history are not included.
+We fitted three demographic models to the simulated data. In addition to
+reinferring the parameters from the simulated model, we fitted two simpler 
+models to the same data. These had fewer parameters, both omitting the size 
+change deeper in time in the ancestral population. This is meant to crudely 
+mimic the scenario that we often face, in which the true history is more 
+complicated than the parameterized model. It also highlights biases in 
+inferences that can arise when features of the true history are not included.
 
@@
 
-\subsection*{Example 2: Inferring human-Neandertal demographic parameters}
+\subsection*{Example 2: Inferring Neanderthal-human demographic parameters}
 
-This example uses publicly available data to infer a demographic model for two
+We used publicly available data to infer a demographic model for two
 modern human populations (a northern European and western African population) 
@@
 from the Neandertal lineage, we are able to recover key parameters such as the
-human-Neandertal divergence time and fraction of ancestry contributed to modern 
-Europeans by Neandertals.
+Neanderthal-human divergence time and the fraction of ancestry contributed to 
+modern Europeans by Neandertals.
 
@@
 1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage 
-sequence of the Vindija Neandertal genome published by \cite{prufer2017high}. 
+sequence of the Vindija Neandertal genome \cite{prufer2017high}. 
 From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
@@
 We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
-were not genotyped in the high-coverage archaic genome sequences, 
-that lacked high-confidence ancestral state assignments, or fell within 10 kb of 
+were not genotyped in four high-coverage archaic genome sequences, 
+lacked high-confidence ancestral state assignments, or fell within 10 kb of 
 exonic or promoter regions. This left us with approximately 960 Mb of 
@@
 The model that we wished to fit is relatively simple but still has many
-parameters ($11$ in the example we present), so we constructed it in stages by
-first fitting marginal one- and two-population models. When our goal is to infer 
+parameters ($11$ in total), so we constructed it in stages by
+first fitting marginal one- and two-population models. When we want to infer 
 a complex demographic history involving multiple populations, it is helpful to 
-start simple, with subsets of the full data, and sequentially add complexity.  
+start simple by modeling subsets of the full data and sequentially adding
+complexity. This procedure gives us initial guesses for parameter values in more
+complex models and may help suggest which features are well-supported by the data.
 Here, we began by fitting one-population models with ancestral and recent size 
-changes for the MSL and GBR populations.  For the MSL model, we allowed three 
-epochs with constant effective size in each epoch. We attributed to GBR a sharp 
-contraction $\sim$60 kya followed by exponential growth to represent the 
+changes for the MSL and GBR populations.  In the MSL model, we allowed three 
+epochs of piecewise-constant effective size. We attributed to GBR two 
+constant-size epochs followed by a sharp 
+contraction and exponential growth beginning $\sim$60 kya to represent the 
 out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
 we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
-and 29 years as the average generation time. YAML files representing the models 
+and 29 years as the average generation time. YAML files encoding the models 
 and parameters in this section can be found in the linked GitHub repository.
@@
 Following this, we added a Neandertal branch and fitted the 
-Neandertal-modern human divergence time, the Neandertal effective size, and the 
-Neandertal-GBR pulse proportion, leaving other parameters fixed. We also 
+Neandertal-human divergence time, the Neandertal effective size, and the 
+Neandertal-GBR pulse proportion, leaving all other parameters fixed. We also 
 fitted the divergence time between the sampled Vindija Neandertal lineage and 
-the Neandertal lineage which admixed with modern non-African humans. Finding 
+the Neandertal lineage which admixed with non-African humans. Finding 
 that this parameter tended to converge to its lower bound (imposed by the 
@@
 value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
-of optimization. Rather than fitting it, we fixed the time of the Neandertal-GBR 
-pulse to 50 kya. Fixing parameters for which good estimates are available in the 
-literature  (see, e.g., \cite{sumer2025earliest}) is often reasonable in the 
-first stages of model optimization, and we expected this value to be especially 
-poorly constrained due to the small pulse proportion and Neandertal sample size. 
-In the second round of optimization on the three-population model, we refit all 
-parameters aside from the two event times mentioned above to refine the 
-estimates.
+of optimization. We also fixed the time of the Neandertal-GBR pulse to 50 kya 
+(at the upper range of values inferrred by \cite{sumer2025earliest}). 
+Fixing parameters for which good estimates are available in the literature is often 
+reasonable in the first stages of model optimization, and we expected this 
+value to be especially poorly constrained due to the small pulse proportion and 
+Neandertal sample size. In the second round of optimization of the 
+three-population model, we refitted all parameters aside from the two times 
+mentioned above to refine the estimates.
 
 After this refitting step, we were left with parameters in good agreement with 
-those inferred by prior authors, with a Neandertal-modern human divergence time 
+those inferred by prior authors, with a Neandertal-human divergence time 
 of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
@@
 likelihood-ratio test (LRT) statistic is overdispersed under composite 
-likelihood, inflating the rate of false positives and rendering its use 
-unjustifiable in this setting. \moments does incorporate a function 
-(\texttt{moments.Godambe.LRT\_adjust}) which computes a multiplicative
-adjustment, intended to restore the expected null distribution, using 
+likelihood, inflating the rate of false positives and preventing its use 
+ in this setting. \moments does incorporate a function 
+(\texttt{moments.Godambe.LRT\_adjust()}) which computes a multiplicative
+adjustment to the LRT, intended to restore the expected null distribution, using 
 first-order moment matching \cite{coffman2016computationally}. This method 
@@
 test by fitting the nested and pulse models to a large number of bootstrap 
-samples and calculating the rate of rejection across replicates-- which has the 
-benefit of producing a distribution of LRT statistics, but is time-consuming
-and computationally expensive.
+samples and calculating the rate of rejection across them-- which has the 
+benefit of producing a distribution of (nonindependent) LRT statistics, but is
+time-consuming and computationally expensive.
 

### 2025-08-26 chapter/paper.tex
final version of revision 1

@@
 \emph{Keywords:} Demographic inference, site frequency spectrum, diffusion
-equation, moments equations
+equation
 
@@
 coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
-kim2017inference}); to scan for historical selective sweeps
-(e.g., \cite{kim2002detecting, nielsen2005genomic}); and to understand how
-selection on quantitative traits impacts the genetic architecture underlying
-those traits (e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS
-has also been used to infer relative mutation rates and how they have changed
-over time \cite{dewitt2021nonparametric}. Each of these analyses typically
-requires partitioning the data in some way -- by functional annotation,
-mutation context, local recombination rate, etc -- so that comparisons can be
-made across SFS from different classes of mutations.
+kim2017inference}); to scan for historical selective sweeps (e.g.,
+\cite{kim2002detecting, nielsen2005genomic}); and to understand how selection
+on quantitative traits impacts the genetic architecture underlying those traits
+(e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS has also been
+used to infer relative mutation rates and how they have changed over time
+\cite{dewitt2021nonparametric}. Each of these analyses typically requires
+partitioning the data in some way -- such as by functional annotation, mutation
+context, local recombination rate -- so that comparisons can be made across SFS
+from different classes of mutations.
 
@@
 \begin{minted}[gobble=4, frame=single,]{yaml}
-    description: split-with-migration model, loosely based on example 2
+    description: split-with-migration example model
     generation_time: 29
@@
 Here, we have not specified a mutation rate. The default behavior sets $u=1$,
-so that $\theta = 4N_eu=4N_e$, where $N_e$ is the ancestral population's
-initial size. With a known mutation rate, the SFS will be properly scaled by
+so that $\theta = 4N_eu=4N_e$, where $N_e$ is the initial size of the ancestral
+population. With a known mutation rate, the SFS will be properly scaled by
 passing the mutation rate as a keyword argument: \texttt{moments.Demes.SFS(g,
@@
 joint SFS. The first example simulates data with known demographic parameters,
-which we try to reinfer using both the original model and simpler
+which we attempt to reinfer using both the original model and simpler
 misspecified models. In the second example, we infer a relatively simple
 Neanderthal-human history using data from two human populations and a
-high-coverage ancient sample from the Neandertal lineage. For both of these
+high-coverage ancient sample from the Neanderthal lineage. For both of these
 examples, we briefly describe data processing and inference setup, emphasizing
@@
 In order to optimize parameters in a \demes model, we need a way to specify
-which parameters should be fitted. For this, we use a separate YAML-formatted file
-to define \texttt{parameters}, each of which points to desired value(s) in the
-demographic model and for which lower and upper bounds may be set. This file
-also specifies any relative \texttt{constraints} between pairs of parameters,
-to ensure the optimization routine only explores valid model space. Below, we
-have included a truncated parameters file -- the full example is available on
-GitHub (URL above).
+which parameters should be fitted. For this, we use a separate YAML-formatted
+file to define \texttt{parameters}, each of which points to desired value(s) in
+the demographic model and for which lower and upper bounds may be set. This
+file also specifies any relative \texttt{constraints} between pairs of
+parameters, to ensure the optimization routine only explores valid model space.
+Below, we have included a truncated parameters file -- the full example is
+available on GitHub (URL above).
 
@@
 
-\subsection*{Example 1: Inferring parameters in a simulated split-with-migration model}
+\subsection*{Example 1: Inferring parameters in a simulated
+split-with-migration model}
 
@@
 modern human populations (a northern European and western African population) 
-and a Neandertal individual. As shown below, even with just a single sample 
-from the Neandertal lineage, we are able to recover key parameters such as the
+and a Neanderthal individual. As shown below, even with just a single sample 
+from the Neanderthal lineage, we are able to recover key parameters such as the
 Neanderthal-human divergence time and the fraction of ancestry contributed to 
-modern Europeans by Neandertals.
+modern Europeans by Neanderthals.
 
@@
 1000 Genomes Project cohort \cite{byrska2022high}, and used the high-coverage
-sequence of the Vindija Neandertal genome \cite{prufer2017high}. 
-From the 1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 
-GBR (British from England and Scotland) sequences as our modern human samples.  
-We filtered out sites that lay outside the 1000 Genomes `strict' mask, 
-were not genotyped in four high-coverage archaic genome sequences, 
-lacked high-confidence ancestral state assignments, or fell within 10 kb of 
-exonic or promoter regions. This left us with approximately 960 Mb of 
-well-characterized and putatively neutrally-evolving sequence.
-We estimated the genome-wide SFS using built-in \moments parsing functions.
-Details concerning data processing (including our liftover of the Vindija 
-sequence to a more recent genome build) can be found in the linked GitHub 
-repository.
+sequence of the Vindija Neanderthal genome \cite{prufer2017high}.  From the
+1000 Genomes cohort we took 85 MSL (Mende from Sierre Leone) and 91 GBR
+(British from England and Scotland) diploid sequences as our modern human
+samples. We filtered out sites that lay outside the 1000 Genomes `strict' mask,
+were not genotyped in four high-coverage archaic genome sequences, lacked
+high-confidence ancestral state assignments, or fell within 10 kb of exonic or
+promoter regions. This left us with approximately 960 Mb of well-characterized
+and putatively neutrally-evolving sequence. We estimated the genome-wide SFS
+using built-in \moments parsing functions. Details concerning data processing
+(including our liftover of the Vindija sequence to a more recent genome build)
+can be found in the linked GitHub repository.
 
 The model that we wished to fit is relatively simple but still has many
-parameters ($11$ in total), so we constructed it in stages by
-first fitting marginal one- and two-population models. When we want to infer 
-a complex demographic history involving multiple populations, it is helpful to 
-start simple by modeling subsets of the full data and sequentially adding
-complexity. This procedure gives us initial guesses for parameter values in more
-complex models and may help suggest which features are well-supported by the data.
+parameters ($11$ in total), so we constructed it in stages by first fitting
+marginal one- and two-population models. When we want to infer a complex
+demographic history involving multiple populations, it is helpful to start
+simple by modeling subsets of the full data and sequentially adding complexity.
+This procedure gives us initial guesses for parameter values in more complex
+models and may help suggest which features are well-supported by the data.
 Here, we began by fitting one-population models with ancestral and recent size
@@
 epochs of piecewise-constant effective size. We attributed to GBR two
-constant-size epochs followed by a sharp 
-contraction and exponential growth beginning $\sim$60 kya to represent the 
-out-of-Africa bottleneck and subsequent rapid population expansion.  Throughout, 
-we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation 
-and 29 years as the average generation time. YAML files encoding the models 
-and parameters in this section can be found in the linked GitHub repository.
+constant-size epochs followed by a sharp contraction and exponential growth
+beginning $\sim$60 kya to represent the out-of-Africa bottleneck and subsequent
+rapid population expansion. Throughout, we used $u=1.5\times 10^{-8}$ as an
+estimate of the mutation rate per generation and 29 years as the average
+generation time. YAML files encoding the models and parameters in this section
+can be found in the linked GitHub repository.
 
-We next stitched these one-population models together and refitted the resulting 
-demographic model to get a two-population model that included continuous 
-symmetric migration between MSL and GBR and an ancestral size expansion. 
-Following this, we added a Neandertal branch and fitted the 
-Neandertal-human divergence time, the Neandertal effective size, and the 
-Neandertal-GBR pulse proportion, leaving all other parameters fixed. We also 
-fitted the divergence time between the sampled Vindija Neandertal lineage and 
-the Neandertal lineage which admixed with non-African humans. Finding 
-that this parameter tended to converge to its lower bound (imposed by the 
-estimated time of death of the Vindija Neandertal, ~55 kya), we fixed it to a 
-value supported by literature (90 kya, \cite{prufer2017high}) in a second round 
-of optimization. We also fixed the time of the Neandertal-GBR pulse to 50 kya 
-(at the upper range of values inferrred by \cite{sumer2025earliest}). 
-Fixing parameters for which good estimates are available in the literature is often 
+We next stitched these one-population models together and refitted the
+resulting demographic model to get a two-population model that included
+continuous symmetric migration between MSL and GBR and an ancestral size
+expansion.  Following this, we added a Neanderthal branch and fitted the
+Neanderthal-human divergence time, the Neanderthal effective size, and the
+Neanderthal-GBR pulse proportion, leaving all other parameters fixed. We also
+fitted the divergence time between the sampled Vindija Neanderthal lineage and
+the Neanderthal lineage which admixed with non-African humans. Finding that
+this parameter tended to converge to its lower bound (imposed by the estimated
+time of death of the Vindija Neanderthal, ~55 kya), we fixed it to a value
+supported by literature (90 kya, \cite{prufer2017high}) in a second round of
+optimization. We also fixed the time of the Neanderthal-GBR pulse to 50 kya (at
+the upper range of values inferrred by \cite{sumer2025earliest}).  Fixing
+parameters for which good estimates are available in the literature is often
 reasonable in the first stages of model optimization, and we expected this
 value to be especially poorly constrained due to the small pulse proportion and
-Neandertal sample size. In the second round of optimization of the 
+Neanderthal sample size. In the second round of optimization of the
 three-population model, we refitted all parameters aside from the two times
@@
 After this refitting step, we were left with parameters in good agreement with
-those inferred by prior authors, with a Neandertal-human divergence time 
-of 547 kya, MSL-GBR split time of 77 kya, and Neandertal-GBR pulse proportion 
-of 2.7 percent. We observe some large, systematic residuals in our best-fit model 
+those inferred by prior authors, with a Neanderthal-human divergence time of
+547 kya, MSL-GBR split time of 77 kya, and Neanderthal-GBR pulse proportion of
+2.7 percent. We observe some large, systematic residuals in our best-fit model
 (Figure~\ref{fig:humans}G-I), suggesting model misspecification that could be
@@
 not perform a formal hypothesis test; the null distribution of the
-likelihood-ratio test (LRT) statistic is overdispersed under composite 
-likelihood, inflating the rate of false positives and preventing its use 
- in this setting. \moments does incorporate a function 
+likelihood-ratio test (LRT) statistic is overdispersed under a composite
+likelihood, inflating the rate of false positives and preventing its use in
+this setting. \moments does incorporate a function
 (\texttt{moments.Godambe.LRT\_adjust()}) which computes a multiplicative
-adjustment to the LRT, intended to restore the expected null distribution, using 
-first-order moment matching \cite{coffman2016computationally}. This method 
-requires bootstrapping over the data. Otherwise, we could conduct a hypothesis 
-test by fitting the nested and pulse models to a large number of bootstrap 
-samples and calculating the rate of rejection across them-- which has the 
-benefit of producing a distribution of (nonindependent) LRT statistics, but is
-time-consuming and computationally expensive.
+adjustment to the LRT, intended to restore the expected null distribution,
+using first-order moment matching \cite{coffman2016computationally}. This
+method requires bootstrapping over the data. Otherwise, we could conduct a
+hypothesis test by fitting the nested and pulse models to a large number of
+bootstrap samples and calculating the rate of rejection across them. This has
+the benefit of producing a distribution of (nonindependent) LRT statistics, but
+is time-consuming and computationally expensive.
 
@@
     \caption{
-        Inference using simulated data. (A) The true demographic model,
-        including an expansion in the ancestral population and symmetric
-        migration between the descendent populations. (B--D) The simulated
-        data, showing both marginal and the joint SFS. (E--G) Residuals
-        between the data SFS (in D) and the predicted SFS from the inferred
-        demographic models in (H--J), resp.  The two misspecified models (H and
-        I) provide worse fits to the data and both overestimate the split time
-        due to disallowing any size change in the ancestral population.
+        \textbf{Inference using simulated data.}
+        (A) The true demographic model, including an expansion in the ancestral
+        population and symmetric migration between the descendent populations.
+        (B--D) The simulated data, showing both marginal and the joint SFS.
+        (E--G) Residuals between the data SFS (in D) and the predicted SFS from
+        the inferred demographic models in (H--J), resp. The two misspecified
+        models (H and I) provide worse fits to the data and both overestimate
+        the split time due to disallowing any size change in the ancestral
+        population.
     }
@@
     \caption{
-        Inference using human and Neandertal data.
-        (A) The best-fit model including the Vindija Neandertal and two human
+        \textbf{Inference using human and Neanderthal data.}
+        (A) The best-fit model including the Vindija Neanderthal and two human
         populations.
